---
title: "analysis-script — OLMo-3 7B"
author: "Zachary Houghton"
date: "2026-03-03"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)
library(plotly)
```

# analysis

## load data
```{r}
up_independently = read_csv('../../Data/olmo-3-7b/Data_up/all_layers_results.csv')
up_subword = read_csv('../../Data/olmo-3-7b/Data_upsubword/all_layers_results.csv')

up_independently = up_independently %>%
  mutate(
    log_freq = log(frequency),
    log_ftp  = log(ftp),
    verb_up  = factor(verb_up)
  )

up_subword = up_subword %>%
  mutate(
    log_freq = log(frequency),
    log_ftp  = log(ftp),
    verb_up  = factor(verb_up)
  )
```

## up independently

### Effect of frequency at the final hidden layer (brms)
```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_up_independently = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_independently %>% filter(layer == 31),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_independently",
  seed = 964
)
```

### Effect of frequency at the first hidden layer (brms)
```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_up_independently_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_independently %>% filter(layer == 0),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_independently_first_layer",
  seed = 964
)
```

### Effect of frequency at the final hidden layer (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_independently_final.rds")) {
  bam_freq_up_independently_final = readRDS("bam_freq_up_independently_final.rds")
} else {
  bam_freq_up_independently_final = bam(
    formula  = logit ~ s(log_freq) + s(verb_up, bs = 're'),
    data     = up_independently %>% filter(layer == 31),
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_independently_final, "bam_freq_up_independently_final.rds")
}

summary(bam_freq_up_independently_final)
```

### Effect of frequency at the first hidden layer (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_independently_first.rds")) {
  bam_freq_up_independently_first = readRDS("bam_freq_up_independently_first.rds")
} else {
  bam_freq_up_independently_first = bam(
    formula  = logit ~ s(log_freq) + s(verb_up, bs = 're'),
    data     = up_independently %>% filter(layer == 0),
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_independently_first, "bam_freq_up_independently_first.rds")
}

summary(bam_freq_up_independently_first)
```

### Effect of frequency interacting with layer (mgcv)
```{r}
set.seed(964)
if (file.exists("model_freq_layer_up_independently.rds")) {
  model_freq_layer_up_independently = readRDS("model_freq_layer_up_independently.rds")
} else {
  model_freq_layer_up_independently = bam(
    formula  = logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
    data     = up_independently,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_independently, "model_freq_layer_up_independently.rds")
}

summary(model_freq_layer_up_independently)
```

## up as a subword

### Effect of frequency at the final hidden layer (brms)
```{r}
model_freq_up_subword = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_subword %>% filter(layer == 31),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_subword",
  seed = 964
)
```

### Effect of frequency at the first hidden layer (brms)
```{r}
model_freq_up_subword_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_subword %>% filter(layer == 0),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_subword_first_layer",
  seed = 964
)
```

### Effect of frequency at the final hidden layer (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_subword_final.rds")) {
  bam_freq_up_subword_final = readRDS("bam_freq_up_subword_final.rds")
} else {
  bam_freq_up_subword_final = bam(
    formula  = logit ~ s(log_freq, k = 10) + s(verb_up, bs = 're'),
    data     = up_subword %>% filter(layer == 31),
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_subword_final, "bam_freq_up_subword_final.rds")
}

summary(bam_freq_up_subword_final)
```

### Effect of frequency at the first hidden layer (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_subword_first.rds")) {
  bam_freq_up_subword_first = readRDS("bam_freq_up_subword_first.rds")
} else {
  bam_freq_up_subword_first = bam(
    formula  = logit ~ s(log_freq, k = 10) + s(verb_up, bs = 're'),
    data     = up_subword %>% filter(layer == 0),
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_subword_first, "bam_freq_up_subword_first.rds")
}

summary(bam_freq_up_subword_first)
```

### Effect of frequency interacting with layer (mgcv)
```{r}
set.seed(964)
if (file.exists("model_freq_layer_up_subword.rds")) {
  model_freq_layer_up_subword = readRDS("model_freq_layer_up_subword.rds")
} else {
  model_freq_layer_up_subword = bam(
    formula  = logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
    data     = up_subword,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_subword, "model_freq_layer_up_subword.rds")
}

summary(model_freq_layer_up_subword)
```

## plots

### final and first layer plots (brms)
```{r}
plot_final_layer = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)

  newdata = data.frame(
    log_freq = log_freq_seq,
    verb_up  = NA
  )

  epred = add_epred_draws(newdata, model, re_formula = NA, ndraws = 500)

  epred %>%
    ggplot(aes(x = log_freq, y = .epred)) +
    stat_lineribbon(.width = c(0.95, 0.80, 0.50), alpha = 0.35) +
    scale_fill_manual(
      values = c("0.95" = "#b3cde3", "0.80" = "#6497b1", "0.50" = "#005b96"),
      labels = c("95%", "80%", "50%")
    ) +
    stat_summary(
      data = data, aes(x = log_freq, y = logit),
      fun      = mean,
      geom     = "point",
      color    = "gray30",
      size     = 1.5,
      alpha    = 0.6
    ) +
    labs(
      title = title,
      x     = "log(frequency)",
      y     = "Predicted logit",
      fill  = "Credible interval"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold", margin = margin(b = 8)),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom",
      legend.key.size  = unit(0.5, "cm")
    )
}

# Final layer
p_indep_final = plot_final_layer(
  model_freq_up_independently,
  up_independently %>% filter(layer == 31),
  "UP independently: frequency effect at final layer (31)"
)

p_subword_final = plot_final_layer(
  model_freq_up_subword,
  up_subword %>% filter(layer == 31),
  "UP as subword: frequency effect at final layer (31)"
)

# First layer
p_indep_first = plot_final_layer(
  model_freq_up_independently_first_layer,
  up_independently %>% filter(layer == 0),
  "UP independently: frequency effect at first layer (0)"
)

p_subword_first = plot_final_layer(
  model_freq_up_subword_first_layer,
  up_subword %>% filter(layer == 0),
  "UP as subword: frequency effect at first layer (0)"
)

# Arrange: rows = layer (first vs final), columns = condition (indep vs subword)
(p_indep_first  | p_subword_first) /
(p_indep_final  | p_subword_final)
```

### final and first layer plots (bam — non-linear)
```{r}
plot_bam_single_layer = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 200)

  newdata = data.frame(
    log_freq = log_freq_seq,
    verb_up  = data$verb_up[1]   # needed for re term; excluded in prediction
  )

  preds = predict(
    model,
    newdata = newdata,
    se.fit  = TRUE,
    exclude = 's(verb_up)'
  )

  newdata = newdata %>%
    mutate(
      fit  = preds$fit,
      se   = preds$se.fit,
      lo95 = fit - 1.96 * se,
      hi95 = fit + 1.96 * se,
      lo80 = fit - 1.28 * se,
      hi80 = fit + 1.28 * se,
      lo50 = fit - 0.67 * se,
      hi50 = fit + 0.67 * se
    )

  ggplot(newdata, aes(x = log_freq, y = fit)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "#b3cde3", alpha = 0.5) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "#6497b1", alpha = 0.5) +
    geom_ribbon(aes(ymin = lo50, ymax = hi50), fill = "#005b96", alpha = 0.5) +
    geom_line(linewidth = 1, color = "#003366") +
    stat_summary(
      data  = data,
      aes(x = log_freq, y = logit),
      fun   = mean,
      geom  = "point",
      color = "gray30",
      size  = 1.5,
      alpha = 0.6
    ) +
    labs(
      title    = title,
      x        = "log(frequency)",
      y        = "Predicted logit",
      subtitle = "Ribbons: 50%, 80%, 95% CI  |  bam() s(log_freq)"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold", margin = margin(b = 4)),
      plot.subtitle    = element_text(size = 9, color = "gray50"),
      panel.grid.minor = element_blank()
    )
}

# Final layer
p_bam_indep_final = plot_bam_single_layer(
  bam_freq_up_independently_final,
  up_independently %>% filter(layer == 31),
  "UP independently: non-linear freq effect at final layer (31)"
)

p_bam_subword_final = plot_bam_single_layer(
  bam_freq_up_subword_final,
  up_subword %>% filter(layer == 31),
  "UP as subword: non-linear freq effect at final layer (31)"
)

# First layer
p_bam_indep_first = plot_bam_single_layer(
  bam_freq_up_independently_first,
  up_independently %>% filter(layer == 0),
  "UP independently: non-linear freq effect at first layer (0)"
)

p_bam_subword_first = plot_bam_single_layer(
  bam_freq_up_subword_first,
  up_subword %>% filter(layer == 0),
  "UP as subword: non-linear freq effect at first layer (0)"
)

# Arrange: rows = layer (first vs final), columns = condition (indep vs subword)
(p_bam_indep_first  | p_bam_subword_first) /
(p_bam_indep_final  | p_bam_subword_final)
```

### brms vs bam comparison (linearity check)
```{r}
# Overlay brms linear fit and bam non-linear fit on the same panel
# to visually assess whether non-linearity matters

plot_brms_vs_bam = function(brms_model, bam_model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 200)

  # --- brms predictions ---
  newdata_brms = data.frame(log_freq = log_freq_seq, verb_up = NA)
  epred = add_epred_draws(newdata_brms, brms_model, re_formula = NA, ndraws = 500)

  brms_ribbon = epred %>%
    group_by(log_freq) %>%
    summarise(
      fit  = mean(.epred),
      lo95 = quantile(.epred, 0.025),
      hi95 = quantile(.epred, 0.975),
      .groups = "drop"
    )

  # --- bam predictions ---
  newdata_bam = data.frame(log_freq = log_freq_seq, verb_up = data$verb_up[1])
  preds_bam   = predict(bam_model, newdata = newdata_bam, se.fit = TRUE,
                        exclude = 's(verb_up)')

  bam_ribbon = data.frame(
    log_freq = log_freq_seq,
    fit      = preds_bam$fit,
    lo95     = preds_bam$fit - 1.96 * preds_bam$se.fit,
    hi95     = preds_bam$fit + 1.96 * preds_bam$se.fit
  )

  # --- raw data points ---
  raw = data %>%
    group_by(log_freq) %>%
    summarise(logit = mean(logit), .groups = "drop")

  ggplot() +
    # brms 95% CI
    geom_ribbon(data = brms_ribbon,
                aes(x = log_freq, ymin = lo95, ymax = hi95, fill = "brms (linear)"),
                alpha = 0.25) +
    geom_line(data = brms_ribbon,
              aes(x = log_freq, y = fit, color = "brms (linear)"),
              linewidth = 1) +
    # bam 95% CI
    geom_ribbon(data = bam_ribbon,
                aes(x = log_freq, ymin = lo95, ymax = hi95, fill = "bam (non-linear)"),
                alpha = 0.25) +
    geom_line(data = bam_ribbon,
              aes(x = log_freq, y = fit, color = "bam (non-linear)"),
              linewidth = 1, linetype = "dashed") +
    # raw means
    geom_point(data = raw,
               aes(x = log_freq, y = logit),
               color = "gray30", size = 1.2, alpha = 0.5) +
    scale_fill_manual(values  = c("brms (linear)" = "#e07b54",
                                   "bam (non-linear)" = "#005b96")) +
    scale_color_manual(values = c("brms (linear)" = "#e07b54",
                                   "bam (non-linear)" = "#005b96")) +
    labs(
      title    = title,
      x        = "log(frequency)",
      y        = "Predicted logit",
      fill     = "Model",
      color    = "Model",
      subtitle = "Points = per-freq mean logit"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold"),
      plot.subtitle    = element_text(size = 9, color = "gray50"),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}

p_compare_indep_final = plot_brms_vs_bam(
  model_freq_up_independently,
  bam_freq_up_independently_final,
  up_independently %>% filter(layer == 31),
  "UP independently: linear vs non-linear at final layer (31)"
)

p_compare_subword_final = plot_brms_vs_bam(
  model_freq_up_subword,
  bam_freq_up_subword_final,
  up_subword %>% filter(layer == 31),
  "UP as subword: linear vs non-linear at final layer (31)"
)

p_compare_indep_first = plot_brms_vs_bam(
  model_freq_up_independently_first_layer,
  bam_freq_up_independently_first,
  up_independently %>% filter(layer == 0),
  "UP independently: linear vs non-linear at first layer (0)"
)

p_compare_subword_first = plot_brms_vs_bam(
  model_freq_up_subword_first_layer,
  bam_freq_up_subword_first,
  up_subword %>% filter(layer == 0),
  "UP as subword: linear vs non-linear at first layer (0)"
)

(p_compare_indep_first  | p_compare_subword_first) /
(p_compare_indep_final  | p_compare_subword_final)
```

### across layers
```{r}
plot_freq_by_layer_mgcv = function(model, data, title, n_layers = 32) {

  layer_seq    = 0:(n_layers - 1)
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo  = quantile(data$log_freq, 0.1)
  log_freq_hi  = quantile(data$log_freq, 0.9)

  # ── 1. Spaghetti: one line per layer across log_freq range ────────────────
  newdata_spaghetti = expand.grid(
    layer    = layer_seq,
    log_freq = log_freq_seq
  ) %>%
    mutate(verb_up = data$verb_up[1])

  preds_spaghetti = predict(
    model,
    newdata = newdata_spaghetti,
    se.fit  = TRUE,
    exclude = 's(verb_up)'
  )

  newdata_spaghetti = newdata_spaghetti %>%
    mutate(
      fit  = preds_spaghetti$fit,
      se   = preds_spaghetti$se.fit,
      lo95 = fit - 1.96 * se,
      hi95 = fit + 1.96 * se
    )

  p_spaghetti = newdata_spaghetti %>%
    ggplot(aes(x = log_freq, y = fit, color = layer, group = factor(layer))) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(
      title = paste0(title, ": effect of log(freq) by layer"),
      x     = "log(frequency)",
      y     = "Predicted logit"
    ) +
    theme_minimal()

  # ── 2. Heatmap: logit surface across freq x layer ─────────────────────────
  p_heatmap = newdata_spaghetti %>%
    ggplot(aes(x = layer, y = log_freq, fill = fit)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(
      title = paste0(title, ": predicted logit surface"),
      x     = "Layer",
      y     = "log(frequency)"
    ) +
    theme_minimal()

  # ── 3. Difference plot: frequency effect (high - low) across layers ────────
  newdata_diff = expand.grid(
    layer      = layer_seq,
    freq_level = c("low", "high")
  ) %>%
    mutate(
      log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
      verb_up  = data$verb_up[1]
    )

  preds_diff = predict(
    model,
    newdata = newdata_diff,
    se.fit  = TRUE,
    exclude = 's(verb_up)'
  )

  newdata_diff = newdata_diff %>%
    mutate(fit = preds_diff$fit, se = preds_diff$se.fit)

  p_diff = newdata_diff %>%
    select(layer, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff,
      lo80    = diff - 1.28 * se_diff,
      hi80    = diff + 1.28 * se_diff
    ) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(title, ": frequency effect (high - low) across layers"),
      x     = "Layer",
      y     = "Difference in predicted logit (90th - 10th pct freq)"
    ) +
    theme_minimal()

  # ── 4. 3D surface (plotly) ─────────────────────────────────────────────────
  fit_matrix = newdata_spaghetti %>%
    select(layer, log_freq, fit) %>%
    pivot_wider(names_from = layer, values_from = fit) %>%
    arrange(log_freq) %>%
    select(-log_freq) %>%
    as.matrix()

  p_3d = plot_ly(
    x = ~layer_seq,
    y = ~log_freq_seq,
    z = ~fit_matrix,
    type       = "surface",
    colorscale = "Viridis",
    showscale  = TRUE
  ) %>%
    layout(
      title = paste0(title, ": 3D predicted logit surface"),
      scene = list(
        xaxis = list(title = "Layer"),
        yaxis = list(title = "log(frequency)"),
        zaxis = list(title = "Predicted logit")
      )
    )

  list(
    spaghetti = p_spaghetti,
    heatmap   = p_heatmap,
    diff      = p_diff,
    surface3d = p_3d
  )
}

# OLMo-3 7B has 32 layers (0–31)
plots_indep = plot_freq_by_layer_mgcv(
  model_freq_layer_up_independently,
  up_independently,
  "UP independently",
  n_layers = 32
)

plots_subword = plot_freq_by_layer_mgcv(
  model_freq_layer_up_subword,
  up_subword,
  "UP as subword",
  n_layers = 32
)

# 2D plots combined
(plots_indep$spaghetti + plots_subword$spaghetti) /
(plots_indep$heatmap   + plots_subword$heatmap)   /
(plots_indep$diff      + plots_subword$diff)

# 3D plots separately (plotly doesn't compose with patchwork)
plots_indep$surface3d
plots_subword$surface3d
```

---

## Predictability (FTP = P(up|V))

FTP analyses require `--corpus-stats-pkl` to have been passed to the pipeline.
Rows with missing FTP are dropped for FTP-specific models.

### Effect of FTP at the final hidden layer (brms)
```{r}
up_indep_final_ftp  = up_independently %>% filter(layer == 31, !is.na(log_ftp))
up_subword_final_ftp = up_subword      %>% filter(layer == 31, !is.na(log_ftp))

model_ftp_up_independently = brm(
  formula = logit ~ log_ftp + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_ftp_up_independently",
  seed    = 964
)
```

### Effect of FTP at the first hidden layer (brms)
```{r}
up_indep_first_ftp  = up_independently %>% filter(layer == 0, !is.na(log_ftp))
up_subword_first_ftp = up_subword      %>% filter(layer == 0, !is.na(log_ftp))

model_ftp_up_independently_first_layer = brm(
  formula = logit ~ log_ftp + (1 | verb_up),
  data    = up_indep_first_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_ftp_up_independently_first_layer",
  seed    = 964
)
```

### Effect of FTP at the final hidden layer (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_ftp_up_independently_final.rds")) {
  bam_ftp_up_independently_final = readRDS("bam_ftp_up_independently_final.rds")
} else {
  bam_ftp_up_independently_final = bam(
    formula  = logit ~ s(log_ftp) + s(verb_up, bs = 're'),
    data     = up_indep_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_ftp_up_independently_final, "bam_ftp_up_independently_final.rds")
}
summary(bam_ftp_up_independently_final)

set.seed(964)
if (file.exists("bam_ftp_up_subword_final.rds")) {
  bam_ftp_up_subword_final = readRDS("bam_ftp_up_subword_final.rds")
} else {
  bam_ftp_up_subword_final = bam(
    formula  = logit ~ s(log_ftp) + s(verb_up, bs = 're'),
    data     = up_subword_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_ftp_up_subword_final, "bam_ftp_up_subword_final.rds")
}
summary(bam_ftp_up_subword_final)
```

### Effect of FTP across layers (mgcv)
```{r}
up_indep_ftp   = up_independently %>% filter(!is.na(log_ftp))
up_subword_ftp = up_subword       %>% filter(!is.na(log_ftp))

set.seed(964)
if (file.exists("model_ftp_layer_up_independently.rds")) {
  model_ftp_layer_up_independently = readRDS("model_ftp_layer_up_independently.rds")
} else {
  model_ftp_layer_up_independently = bam(
    formula  = logit ~ te(log_ftp, layer) + s(verb_up, bs = 're'),
    data     = up_indep_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_ftp_layer_up_independently, "model_ftp_layer_up_independently.rds")
}
summary(model_ftp_layer_up_independently)

set.seed(964)
if (file.exists("model_ftp_layer_up_subword.rds")) {
  model_ftp_layer_up_subword = readRDS("model_ftp_layer_up_subword.rds")
} else {
  model_ftp_layer_up_subword = bam(
    formula  = logit ~ te(log_ftp, layer) + s(verb_up, bs = 're'),
    data     = up_subword_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_ftp_layer_up_subword, "model_ftp_layer_up_subword.rds")
}
summary(model_ftp_layer_up_subword)
```

### FTP across layers — plots
```{r}
# Reuse plot_freq_by_layer_mgcv with log_ftp substituted for log_freq
plot_ftp_by_layer_mgcv = function(model, data, title, n_layers = 32) {
  layer_seq   = 0:(n_layers - 1)
  ftp_seq     = seq(min(data$log_ftp, na.rm = TRUE), max(data$log_ftp, na.rm = TRUE), length.out = 50)
  ftp_lo      = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi      = quantile(data$log_ftp, 0.9, na.rm = TRUE)

  newdata_spaghetti = expand.grid(layer = layer_seq, log_ftp = ftp_seq) %>%
    mutate(verb_up = data$verb_up[1])

  preds = predict(model, newdata = newdata_spaghetti, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_spaghetti$fit = preds$fit

  p_spaghetti = newdata_spaghetti %>%
    ggplot(aes(x = log_ftp, y = fit, color = layer, group = factor(layer))) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(title = paste0(title, ": FTP effect by layer"), x = "log(FTP)", y = "Predicted logit") +
    theme_minimal()

  p_heatmap = newdata_spaghetti %>%
    ggplot(aes(x = layer, y = log_ftp, fill = fit)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = paste0(title, ": predicted logit surface (FTP)"), x = "Layer", y = "log(FTP)") +
    theme_minimal()

  newdata_diff = expand.grid(layer = layer_seq, ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi), verb_up = data$verb_up[1])

  preds_diff = predict(model, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit
  newdata_diff$se  = preds_diff$se.fit

  p_diff = newdata_diff %>%
    select(layer, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff,
      lo80    = diff - 1.28 * se_diff,
      hi80    = diff + 1.28 * se_diff
    ) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "darkorange", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "darkorange", alpha = 0.3) +
    geom_line(linewidth = 1, color = "darkorange4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(title, ": FTP effect (high - low) across layers"),
      x = "Layer", y = "Difference in predicted logit (90th - 10th pct FTP)"
    ) +
    theme_minimal()

  list(spaghetti = p_spaghetti, heatmap = p_heatmap, diff = p_diff)
}

plots_ftp_indep   = plot_ftp_by_layer_mgcv(model_ftp_layer_up_independently, up_indep_ftp,   "UP independently", n_layers = 32)
plots_ftp_subword = plot_ftp_by_layer_mgcv(model_ftp_layer_up_subword,       up_subword_ftp, "UP as subword",    n_layers = 32)

(plots_ftp_indep$spaghetti + plots_ftp_subword$spaghetti) /
(plots_ftp_indep$heatmap   + plots_ftp_subword$heatmap)   /
(plots_ftp_indep$diff      + plots_ftp_subword$diff)
```

---

## Joint frequency + predictability

### Additive freq + FTP at the final layer (brms)
```{r}
model_freq_ftp_up_independently = brm(
  formula = logit ~ log_freq + log_ftp + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_ftp_up_independently",
  seed    = 964
)

model_freq_ftp_up_subword = brm(
  formula = logit ~ log_freq + log_ftp + (1 | verb_up),
  data    = up_subword_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_ftp_up_subword",
  seed    = 964
)
```

### Joint freq × FTP surface at final layer (bam)
```{r}
set.seed(964)
if (file.exists("bam_joint_up_independently_final.rds")) {
  bam_joint_up_independently_final = readRDS("bam_joint_up_independently_final.rds")
} else {
  bam_joint_up_independently_final = bam(
    formula  = logit ~ te(log_freq, log_ftp) + s(verb_up, bs = 're'),
    data     = up_indep_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_joint_up_independently_final, "bam_joint_up_independently_final.rds")
}
summary(bam_joint_up_independently_final)

set.seed(964)
if (file.exists("bam_joint_up_subword_final.rds")) {
  bam_joint_up_subword_final = readRDS("bam_joint_up_subword_final.rds")
} else {
  bam_joint_up_subword_final = bam(
    formula  = logit ~ te(log_freq, log_ftp) + s(verb_up, bs = 're'),
    data     = up_subword_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_joint_up_subword_final, "bam_joint_up_subword_final.rds")
}
summary(bam_joint_up_subword_final)
```

### Joint freq × FTP surface — plot
```{r}
plot_joint_surface = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq, na.rm = TRUE), max(data$log_freq, na.rm = TRUE), length.out = 50)
  log_ftp_seq  = seq(min(data$log_ftp,  na.rm = TRUE), max(data$log_ftp,  na.rm = TRUE), length.out = 50)

  newdata = expand.grid(log_freq = log_freq_seq, log_ftp = log_ftp_seq) %>%
    mutate(verb_up = data$verb_up[1])

  preds = predict(model, newdata = newdata, se.fit = FALSE, exclude = 's(verb_up)')
  newdata$fit = preds

  p_heat = newdata %>%
    ggplot(aes(x = log_freq, y = log_ftp, fill = fit)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = title, x = "log(frequency)", y = "log(FTP)") +
    theme_minimal()

  fit_matrix = newdata %>%
    select(log_freq, log_ftp, fit) %>%
    pivot_wider(names_from = log_freq, values_from = fit) %>%
    select(-log_ftp) %>%
    as.matrix()

  p_3d = plot_ly(
    x = ~log_freq_seq, y = ~log_ftp_seq, z = ~fit_matrix,
    type = "surface", colorscale = "Viridis"
  ) %>%
    layout(
      title = paste0(title, " — 3D"),
      scene = list(
        xaxis = list(title = "log(freq)"),
        yaxis = list(title = "log(FTP)"),
        zaxis = list(title = "Predicted logit")
      )
    )

  list(heatmap = p_heat, surface3d = p_3d)
}

joint_indep   = plot_joint_surface(bam_joint_up_independently_final, up_indep_final_ftp,  "Joint freq×FTP: UP independently (final layer)")
joint_subword = plot_joint_surface(bam_joint_up_subword_final,       up_subword_final_ftp, "Joint freq×FTP: UP as subword (final layer)")

joint_indep$heatmap + joint_subword$heatmap

joint_indep$surface3d
joint_subword$surface3d
```

### Freq and FTP effects across layers (additive bam)
```{r}
set.seed(964)
if (file.exists("model_freq_ftp_layer_up_independently.rds")) {
  model_freq_ftp_layer_up_independently = readRDS("model_freq_ftp_layer_up_independently.rds")
} else {
  model_freq_ftp_layer_up_independently = bam(
    formula  = logit ~ te(log_freq, layer) + te(log_ftp, layer) + s(verb_up, bs = 're'),
    data     = up_indep_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_ftp_layer_up_independently, "model_freq_ftp_layer_up_independently.rds")
}
summary(model_freq_ftp_layer_up_independently)

set.seed(964)
if (file.exists("model_freq_ftp_layer_up_subword.rds")) {
  model_freq_ftp_layer_up_subword = readRDS("model_freq_ftp_layer_up_subword.rds")
} else {
  model_freq_ftp_layer_up_subword = bam(
    formula  = logit ~ te(log_freq, layer) + te(log_ftp, layer) + s(verb_up, bs = 're'),
    data     = up_subword_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_ftp_layer_up_subword, "model_freq_ftp_layer_up_subword.rds")
}
summary(model_freq_ftp_layer_up_subword)
```
