---
title: "Unified Analysis — LLM Phrasal Compositionality"
author: "Zachary Houghton"
date: "`r Sys.Date()`"
output: html_document
params:
  source: "olmo"   # "olmo" | "babylm" | "whisper"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)
library(plotly)
library(jsonlite)
```

# Configuration

Set `params$source` to `"olmo"`, `"babylm"`, or `"whisper"` in the YAML header
(or via `rmarkdown::render(..., params = list(source = "babylm"))`).

Fitted model objects (brms and bam) are cached in `../model_cache/{source}/`
which is gitignored so large files are not uploaded to GitHub.

```{r config}
SOURCE    = params$source
CACHE_DIR = file.path("../model_cache", SOURCE)
dir.create(CACHE_DIR, recursive = TRUE, showWarnings = FALSE)

has_ftp         = SOURCE %in% c("olmo", "babylm")
has_component   = SOURCE == "whisper"
has_multi_model = SOURCE == "babylm"

priors = c(prior(normal(0, 1), class = 'b'))

# Helper: full path for bam .rds cache files
rds_path = function(name) file.path(CACHE_DIR, paste0(name, ".rds"))

# Helper: full path prefix for brms file= argument (brms appends .rds itself)
brms_path = function(name) file.path(CACHE_DIR, name)

# Helper: load bam from cache or fit and save
cache_bam = function(name, expr_fn) {
  p = rds_path(name)
  if (file.exists(p)) readRDS(p) else { m = expr_fn(); saveRDS(m, p); m }
}
```

---

# Data loading

```{r load-olmo, eval = params$source == "olmo"}
up_independently = read_csv('../Data/olmo-3-7b/Data_up/all_layers_results.csv') %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))

up_subword = read_csv('../Data/olmo-3-7b/Data_upsubword/all_layers_results.csv') %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))

FINAL_LAYER = 31
N_LAYERS    = 32

up_indep_final = up_independently %>% filter(layer == FINAL_LAYER)
up_sub_final   = up_subword       %>% filter(layer == FINAL_LAYER)
up_indep_first = up_independently %>% filter(layer == 0)
up_sub_first   = up_subword       %>% filter(layer == 0)
```

```{r load-babylm, eval = params$source == "babylm"}
FINAL_LAYER_MAP = list("opt-125m" = 11L, "opt-350m" = 23L, "opt-1.3b" = 23L)

load_babylm_model = function(tag) {
  ui = read_csv(paste0('../Data/babylm/', tag, '/Data_up/all_layers_results.csv')) %>%
    mutate(model = tag, log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))
  us = read_csv(paste0('../Data/babylm/', tag, '/Data_upsubword/all_layers_results.csv')) %>%
    mutate(model = tag, log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))
  list(ui = ui, us = us)
}

raw = lapply(c("opt-125m", "opt-350m", "opt-1.3b"), load_babylm_model)
names(raw) = c("opt-125m", "opt-350m", "opt-1.3b")

up_independently = map_dfr(raw, "ui") %>%
  mutate(model = factor(model, levels = c("opt-125m", "opt-350m", "opt-1.3b"))) %>%
  group_by(model) %>% mutate(layer_norm = layer / max(layer)) %>% ungroup()

up_subword = map_dfr(raw, "us") %>%
  mutate(model = factor(model, levels = c("opt-125m", "opt-350m", "opt-1.3b"))) %>%
  group_by(model) %>% mutate(layer_norm = layer / max(layer)) %>% ungroup()

up_indep_final = up_independently %>%
  group_by(model) %>%
  filter(layer == FINAL_LAYER_MAP[[as.character(model[1])]]) %>% ungroup()

up_sub_final = up_subword %>%
  group_by(model) %>%
  filter(layer == FINAL_LAYER_MAP[[as.character(model[1])]]) %>% ungroup()

up_indep_first = up_independently %>% filter(layer == 0)
up_sub_first   = up_subword       %>% filter(layer == 0)
```

```{r load-whisper, eval = params$source == "whisper"}
encoder = read_csv('../Data/whisper/encoder/all_layers_results.csv') %>%
  mutate(component = "encoder", log_freq = log(frequency), verb_up = factor(verb_up))

decoder = read_csv('../Data/whisper/decoder/all_layers_results.csv') %>%
  mutate(component = "decoder", log_freq = log(frequency), verb_up = factor(verb_up))

whisper_all = bind_rows(encoder, decoder) %>%
  mutate(component = factor(component, levels = c("encoder", "decoder")))

FINAL_LAYER   = 11
N_LAYERS      = 12

whisper_final = whisper_all %>% filter(layer == FINAL_LAYER)
whisper_first = whisper_all %>% filter(layer == 0)
enc_final     = encoder     %>% filter(layer == FINAL_LAYER)
dec_final     = decoder     %>% filter(layer == FINAL_LAYER)
```

---

# OLMo-3 7B Analysis

`r if (params$source != "olmo") "*(Not running — set params$source = 'olmo')*"`

## UP independently

### Effect of frequency at the final layer (brms)
```{r olmo-brms-freq-indep-final, eval = params$source == "olmo"}
model_freq_up_independently = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data    = up_indep_final,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_independently")
)
summary(model_freq_up_independently)
```

### Effect of frequency at the first layer (brms)
```{r olmo-brms-freq-indep-first, eval = params$source == "olmo"}
model_freq_up_independently_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data    = up_indep_first,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_independently_first_layer")
)
summary(model_freq_up_independently_first_layer)
```

### Effect of frequency at the final layer (bam — non-linear)
```{r olmo-bam-freq-indep-final, eval = params$source == "olmo"}
set.seed(964)
bam_freq_up_independently_final = cache_bam("bam_freq_up_independently_final", function() {
  bam(logit ~ s(log_freq) + s(verb_up, bs = 're'),
      data = up_indep_final, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_independently_final)
```

### Effect of frequency at the first layer (bam — non-linear)
```{r olmo-bam-freq-indep-first, eval = params$source == "olmo"}
set.seed(964)
bam_freq_up_independently_first = cache_bam("bam_freq_up_independently_first", function() {
  bam(logit ~ s(log_freq) + s(verb_up, bs = 're'),
      data = up_indep_first, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_independently_first)
```

### Effect of frequency interacting with layer (mgcv)
```{r olmo-bam-freq-layer-indep, eval = params$source == "olmo"}
set.seed(964)
model_freq_layer_up_independently = cache_bam("model_freq_layer_up_independently", function() {
  bam(logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
      data = up_independently, method = 'fREML', discrete = TRUE)
})
summary(model_freq_layer_up_independently)
```

## UP as a subword

### Effect of frequency at the final layer (brms)
```{r olmo-brms-freq-sub-final, eval = params$source == "olmo"}
model_freq_up_subword = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data    = up_sub_final,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_subword")
)
summary(model_freq_up_subword)
```

### Effect of frequency at the first layer (brms)
```{r olmo-brms-freq-sub-first, eval = params$source == "olmo"}
model_freq_up_subword_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data    = up_sub_first,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_subword_first_layer")
)
summary(model_freq_up_subword_first_layer)
```

### Effect of frequency at the final layer (bam — non-linear)
```{r olmo-bam-freq-sub-final, eval = params$source == "olmo"}
set.seed(964)
bam_freq_up_subword_final = cache_bam("bam_freq_up_subword_final", function() {
  bam(logit ~ s(log_freq, k = 10) + s(verb_up, bs = 're'),
      data = up_sub_final, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_subword_final)
```

### Effect of frequency at the first layer (bam — non-linear)
```{r olmo-bam-freq-sub-first, eval = params$source == "olmo"}
set.seed(964)
bam_freq_up_subword_first = cache_bam("bam_freq_up_subword_first", function() {
  bam(logit ~ s(log_freq, k = 10) + s(verb_up, bs = 're'),
      data = up_sub_first, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_subword_first)
```

### Effect of frequency interacting with layer (mgcv)
```{r olmo-bam-freq-layer-sub, eval = params$source == "olmo"}
set.seed(964)
model_freq_layer_up_subword = cache_bam("model_freq_layer_up_subword", function() {
  bam(logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
      data = up_subword, method = 'fREML', discrete = TRUE)
})
summary(model_freq_layer_up_subword)
```

## OLMo Plots

### Plot helper functions
```{r olmo-plot-helpers, eval = params$source == "olmo"}
plot_final_layer = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)
  newdata = data.frame(log_freq = log_freq_seq, verb_up = NA)
  epred = add_epred_draws(newdata, model, re_formula = NA, ndraws = 500)
  epred %>%
    ggplot(aes(x = log_freq, y = .epred)) +
    stat_lineribbon(.width = c(0.95, 0.80, 0.50), alpha = 0.35) +
    scale_fill_manual(
      values = c("0.95" = "#b3cde3", "0.80" = "#6497b1", "0.50" = "#005b96"),
      labels = c("95%", "80%", "50%")
    ) +
    stat_summary(data = data, aes(x = log_freq, y = logit),
                 fun = mean, geom = "point", color = "gray30", size = 1.5, alpha = 0.6) +
    labs(title = title, x = "log(frequency)", y = "Predicted logit", fill = "Credible interval") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 11, face = "bold", margin = margin(b = 8)),
          panel.grid.minor = element_blank(), legend.position = "bottom",
          legend.key.size = unit(0.5, "cm"))
}

plot_bam_single_layer = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 200)
  newdata = data.frame(log_freq = log_freq_seq, verb_up = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = TRUE, exclude = 's(verb_up)')
  newdata = newdata %>%
    mutate(fit = preds$fit, se = preds$se.fit,
           lo95 = fit - 1.96 * se, hi95 = fit + 1.96 * se,
           lo80 = fit - 1.28 * se, hi80 = fit + 1.28 * se,
           lo50 = fit - 0.67 * se, hi50 = fit + 0.67 * se)
  ggplot(newdata, aes(x = log_freq, y = fit)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "#b3cde3", alpha = 0.5) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "#6497b1", alpha = 0.5) +
    geom_ribbon(aes(ymin = lo50, ymax = hi50), fill = "#005b96", alpha = 0.5) +
    geom_line(linewidth = 1, color = "#003366") +
    stat_summary(data = data, aes(x = log_freq, y = logit),
                 fun = mean, geom = "point", color = "gray30", size = 1.5, alpha = 0.6) +
    labs(title = title, x = "log(frequency)", y = "Predicted logit",
         subtitle = "Ribbons: 50%, 80%, 95% CI  |  bam() s(log_freq)") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 11, face = "bold", margin = margin(b = 4)),
          plot.subtitle = element_text(size = 9, color = "gray50"),
          panel.grid.minor = element_blank())
}

plot_brms_vs_bam = function(brms_model, bam_model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 200)
  newdata_brms = data.frame(log_freq = log_freq_seq, verb_up = NA)
  epred = add_epred_draws(newdata_brms, brms_model, re_formula = NA, ndraws = 500)
  brms_ribbon = epred %>% group_by(log_freq) %>%
    summarise(fit = mean(.epred), lo95 = quantile(.epred, 0.025),
              hi95 = quantile(.epred, 0.975), .groups = "drop")
  newdata_bam = data.frame(log_freq = log_freq_seq, verb_up = data$verb_up[1])
  preds_bam   = predict(bam_model, newdata = newdata_bam, se.fit = TRUE, exclude = 's(verb_up)')
  bam_ribbon  = data.frame(log_freq = log_freq_seq, fit = preds_bam$fit,
                            lo95 = preds_bam$fit - 1.96 * preds_bam$se.fit,
                            hi95 = preds_bam$fit + 1.96 * preds_bam$se.fit)
  raw = data %>% group_by(log_freq) %>% summarise(logit = mean(logit), .groups = "drop")
  ggplot() +
    geom_ribbon(data = brms_ribbon, aes(x = log_freq, ymin = lo95, ymax = hi95,
                                         fill = "brms (linear)"), alpha = 0.25) +
    geom_line(data = brms_ribbon, aes(x = log_freq, y = fit, color = "brms (linear)"),
              linewidth = 1) +
    geom_ribbon(data = bam_ribbon, aes(x = log_freq, ymin = lo95, ymax = hi95,
                                        fill = "bam (non-linear)"), alpha = 0.25) +
    geom_line(data = bam_ribbon, aes(x = log_freq, y = fit, color = "bam (non-linear)"),
              linewidth = 1, linetype = "dashed") +
    geom_point(data = raw, aes(x = log_freq, y = logit), color = "gray30", size = 1.2, alpha = 0.5) +
    scale_fill_manual(values  = c("brms (linear)" = "#e07b54", "bam (non-linear)" = "#005b96")) +
    scale_color_manual(values = c("brms (linear)" = "#e07b54", "bam (non-linear)" = "#005b96")) +
    labs(title = title, x = "log(frequency)", y = "Predicted logit",
         fill = "Model", color = "Model", subtitle = "Points = per-freq mean logit") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 11, face = "bold"),
          plot.subtitle = element_text(size = 9, color = "gray50"),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}

plot_freq_by_layer_mgcv = function(model, data, title, n_layers = 32) {
  layer_seq    = 0:(n_layers - 1)
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo  = quantile(data$log_freq, 0.1)
  log_freq_hi  = quantile(data$log_freq, 0.9)
  newdata_sp = expand.grid(layer = layer_seq, log_freq = log_freq_seq) %>%
    mutate(verb_up = data$verb_up[1])
  preds_sp = predict(model, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp = newdata_sp %>%
    mutate(fit = preds_sp$fit, se = preds_sp$se.fit,
           lo95 = fit - 1.96 * se, hi95 = fit + 1.96 * se)
  p_spaghetti = newdata_sp %>%
    ggplot(aes(x = log_freq, y = fit, color = layer, group = factor(layer))) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(title = paste0(title, ": effect of log(freq) by layer"),
         x = "log(frequency)", y = "Predicted logit") +
    theme_minimal()
  p_heatmap = newdata_sp %>%
    ggplot(aes(x = layer, y = log_freq, fill = fit)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = paste0(title, ": predicted logit surface"),
         x = "Layer", y = "log(frequency)") +
    theme_minimal()
  newdata_diff = expand.grid(layer = layer_seq, freq_level = c("low", "high")) %>%
    mutate(log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
           verb_up  = data$verb_up[1])
  preds_diff = predict(model, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff = newdata_diff %>% mutate(fit = preds_diff$fit, se = preds_diff$se.fit)
  p_diff = newdata_diff %>%
    select(layer, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(diff = fit_high - fit_low, se_diff = sqrt(se_high^2 + se_low^2),
           lo95 = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80 = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(title, ": frequency effect (high − low) across layers"),
         x = "Layer", y = "Difference in predicted logit (90th − 10th pct freq)") +
    theme_minimal()
  fit_matrix = newdata_sp %>% select(layer, log_freq, fit) %>%
    pivot_wider(names_from = layer, values_from = fit) %>%
    arrange(log_freq) %>% select(-log_freq) %>% as.matrix()
  p_3d = plot_ly(x = ~layer_seq, y = ~log_freq_seq, z = ~fit_matrix,
                 type = "surface", colorscale = "Viridis", showscale = TRUE) %>%
    layout(title = paste0(title, ": 3D predicted logit surface"),
           scene = list(xaxis = list(title = "Layer"),
                        yaxis = list(title = "log(frequency)"),
                        zaxis = list(title = "Predicted logit")))
  list(spaghetti = p_spaghetti, heatmap = p_heatmap, diff = p_diff, surface3d = p_3d)
}

plot_ftp_by_layer_mgcv = function(model, data, title, n_layers = 32) {
  layer_seq = 0:(n_layers - 1)
  ftp_seq   = seq(min(data$log_ftp, na.rm = TRUE), max(data$log_ftp, na.rm = TRUE), length.out = 50)
  ftp_lo    = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi    = quantile(data$log_ftp, 0.9, na.rm = TRUE)
  newdata_sp = expand.grid(layer = layer_seq, log_ftp = ftp_seq) %>%
    mutate(verb_up = data$verb_up[1])
  preds = predict(model, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp$fit = preds$fit
  p_spaghetti = newdata_sp %>%
    ggplot(aes(x = log_ftp, y = fit, color = layer, group = factor(layer))) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(title = paste0(title, ": FTP effect by layer"), x = "log(FTP)", y = "Predicted logit") +
    theme_minimal()
  p_heatmap = newdata_sp %>%
    ggplot(aes(x = layer, y = log_ftp, fill = fit)) + geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = paste0(title, ": predicted logit surface (FTP)"),
         x = "Layer", y = "log(FTP)") + theme_minimal()
  newdata_diff = expand.grid(layer = layer_seq, ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi), verb_up = data$verb_up[1])
  preds_diff = predict(model, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit; newdata_diff$se = preds_diff$se.fit
  p_diff = newdata_diff %>%
    select(layer, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(diff = fit_high - fit_low, se_diff = sqrt(se_high^2 + se_low^2),
           lo95 = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80 = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "darkorange", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "darkorange", alpha = 0.3) +
    geom_line(linewidth = 1, color = "darkorange4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(title, ": FTP effect (high − low) across layers"),
         x = "Layer", y = "Difference in predicted logit (90th − 10th pct FTP)") +
    theme_minimal()
  list(spaghetti = p_spaghetti, heatmap = p_heatmap, diff = p_diff)
}

plot_joint_surface = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq, na.rm = TRUE), max(data$log_freq, na.rm = TRUE), length.out = 50)
  log_ftp_seq  = seq(min(data$log_ftp,  na.rm = TRUE), max(data$log_ftp,  na.rm = TRUE), length.out = 50)
  newdata = expand.grid(log_freq = log_freq_seq, log_ftp = log_ftp_seq) %>%
    mutate(verb_up = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = FALSE, exclude = 's(verb_up)')
  newdata$fit = preds
  p_heat = newdata %>%
    ggplot(aes(x = log_freq, y = log_ftp, fill = fit)) + geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = title, x = "log(frequency)", y = "log(FTP)") + theme_minimal()
  fit_matrix = newdata %>% select(log_freq, log_ftp, fit) %>%
    pivot_wider(names_from = log_freq, values_from = fit) %>%
    select(-log_ftp) %>% as.matrix()
  p_3d = plot_ly(x = ~log_freq_seq, y = ~log_ftp_seq, z = ~fit_matrix,
                 type = "surface", colorscale = "Viridis") %>%
    layout(title = paste0(title, " — 3D"),
           scene = list(xaxis = list(title = "log(freq)"),
                        yaxis = list(title = "log(FTP)"),
                        zaxis = list(title = "Predicted logit")))
  list(heatmap = p_heat, surface3d = p_3d)
}
```

### Frequency at first/final layer — brms plots
```{r olmo-plot-brms-freq, eval = params$source == "olmo"}
p_indep_final = plot_final_layer(model_freq_up_independently,
  up_indep_final, "UP independently: frequency effect at final layer (31)")
p_sub_final   = plot_final_layer(model_freq_up_subword,
  up_sub_final,   "UP as subword: frequency effect at final layer (31)")
p_indep_first = plot_final_layer(model_freq_up_independently_first_layer,
  up_indep_first, "UP independently: frequency effect at first layer (0)")
p_sub_first   = plot_final_layer(model_freq_up_subword_first_layer,
  up_sub_first,   "UP as subword: frequency effect at first layer (0)")
(p_indep_first | p_sub_first) / (p_indep_final | p_sub_final)
```

### Frequency at first/final layer — bam plots
```{r olmo-plot-bam-freq, eval = params$source == "olmo"}
p_bam_indep_final = plot_bam_single_layer(bam_freq_up_independently_final,
  up_indep_final, "UP independently: non-linear freq effect at final layer (31)")
p_bam_sub_final   = plot_bam_single_layer(bam_freq_up_subword_final,
  up_sub_final,   "UP as subword: non-linear freq effect at final layer (31)")
p_bam_indep_first = plot_bam_single_layer(bam_freq_up_independently_first,
  up_indep_first, "UP independently: non-linear freq effect at first layer (0)")
p_bam_sub_first   = plot_bam_single_layer(bam_freq_up_subword_first,
  up_sub_first,   "UP as subword: non-linear freq effect at first layer (0)")
(p_bam_indep_first | p_bam_sub_first) / (p_bam_indep_final | p_bam_sub_final)
```

### brms vs bam comparison (linearity check)
```{r olmo-plot-brms-vs-bam, eval = params$source == "olmo"}
p_cmp_indep_final = plot_brms_vs_bam(model_freq_up_independently,
  bam_freq_up_independently_final, up_indep_final,
  "UP independently: linear vs non-linear at final layer (31)")
p_cmp_sub_final = plot_brms_vs_bam(model_freq_up_subword,
  bam_freq_up_subword_final, up_sub_final,
  "UP as subword: linear vs non-linear at final layer (31)")
p_cmp_indep_first = plot_brms_vs_bam(model_freq_up_independently_first_layer,
  bam_freq_up_independently_first, up_indep_first,
  "UP independently: linear vs non-linear at first layer (0)")
p_cmp_sub_first = plot_brms_vs_bam(model_freq_up_subword_first_layer,
  bam_freq_up_subword_first, up_sub_first,
  "UP as subword: linear vs non-linear at first layer (0)")
(p_cmp_indep_first | p_cmp_sub_first) / (p_cmp_indep_final | p_cmp_sub_final)
```

### Frequency across all layers — plots
```{r olmo-plot-freq-layers, eval = params$source == "olmo"}
plots_indep   = plot_freq_by_layer_mgcv(model_freq_layer_up_independently,
                                        up_independently, "UP independently", n_layers = 32)
plots_subword = plot_freq_by_layer_mgcv(model_freq_layer_up_subword,
                                        up_subword, "UP as subword", n_layers = 32)
(plots_indep$spaghetti + plots_subword$spaghetti) /
(plots_indep$heatmap   + plots_subword$heatmap)   /
(plots_indep$diff      + plots_subword$diff)
plots_indep$surface3d
plots_subword$surface3d
```

---

## OLMo — Predictability (FTP = P(up|V))

### Effect of FTP at the final layer (brms)
```{r olmo-brms-ftp-final, eval = params$source == "olmo"}
up_indep_final_ftp  = up_independently %>% filter(layer == 31, !is.na(log_ftp))
up_sub_final_ftp    = up_subword       %>% filter(layer == 31, !is.na(log_ftp))

model_ftp_up_independently = brm(
  formula = logit ~ log_ftp + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_ftp_up_independently")
)

model_ftp_up_subword = brm(
  formula = logit ~ log_ftp + (1 | verb_up),
  data    = up_sub_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_ftp_up_subword")
)
summary(model_ftp_up_independently)
summary(model_ftp_up_subword)
```

### Effect of FTP at the first layer (brms)
```{r olmo-brms-ftp-first, eval = params$source == "olmo"}
up_indep_first_ftp = up_independently %>% filter(layer == 0, !is.na(log_ftp))
up_sub_first_ftp   = up_subword       %>% filter(layer == 0, !is.na(log_ftp))

model_ftp_up_independently_first_layer = brm(
  formula = logit ~ log_ftp + (1 | verb_up),
  data    = up_indep_first_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_ftp_up_independently_first_layer")
)
summary(model_ftp_up_independently_first_layer)
```

### Effect of FTP at the final layer (bam — non-linear)
```{r olmo-bam-ftp-final, eval = params$source == "olmo"}
set.seed(964)
bam_ftp_up_independently_final = cache_bam("bam_ftp_up_independently_final", function() {
  bam(logit ~ s(log_ftp) + s(verb_up, bs = 're'),
      data = up_indep_final_ftp, method = 'fREML', discrete = TRUE)
})
bam_ftp_up_subword_final = cache_bam("bam_ftp_up_subword_final", function() {
  bam(logit ~ s(log_ftp) + s(verb_up, bs = 're'),
      data = up_sub_final_ftp, method = 'fREML', discrete = TRUE)
})
summary(bam_ftp_up_independently_final)
summary(bam_ftp_up_subword_final)
```

### Effect of FTP across layers (mgcv)
```{r olmo-bam-ftp-layers, eval = params$source == "olmo"}
up_indep_ftp   = up_independently %>% filter(!is.na(log_ftp))
up_subword_ftp = up_subword       %>% filter(!is.na(log_ftp))

set.seed(964)
model_ftp_layer_up_independently = cache_bam("model_ftp_layer_up_independently", function() {
  bam(logit ~ te(log_ftp, layer) + s(verb_up, bs = 're'),
      data = up_indep_ftp, method = 'fREML', discrete = TRUE)
})
model_ftp_layer_up_subword = cache_bam("model_ftp_layer_up_subword", function() {
  bam(logit ~ te(log_ftp, layer) + s(verb_up, bs = 're'),
      data = up_subword_ftp, method = 'fREML', discrete = TRUE)
})
summary(model_ftp_layer_up_independently)
summary(model_ftp_layer_up_subword)
```

### FTP across layers — plots
```{r olmo-plot-ftp-layers, eval = params$source == "olmo"}
plots_ftp_indep   = plot_ftp_by_layer_mgcv(model_ftp_layer_up_independently,
                                            up_indep_ftp, "UP independently", n_layers = 32)
plots_ftp_subword = plot_ftp_by_layer_mgcv(model_ftp_layer_up_subword,
                                            up_subword_ftp, "UP as subword", n_layers = 32)
(plots_ftp_indep$spaghetti + plots_ftp_subword$spaghetti) /
(plots_ftp_indep$heatmap   + plots_ftp_subword$heatmap)   /
(plots_ftp_indep$diff      + plots_ftp_subword$diff)
```

---

## OLMo — Joint frequency + predictability

### Additive freq + FTP at the final layer (brms)
```{r olmo-brms-joint, eval = params$source == "olmo"}
model_freq_ftp_up_independently = brm(
  formula = logit ~ log_freq + log_ftp + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_ftp_up_independently")
)
model_freq_ftp_up_subword = brm(
  formula = logit ~ log_freq + log_ftp + (1 | verb_up),
  data    = up_sub_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_ftp_up_subword")
)
summary(model_freq_ftp_up_independently)
summary(model_freq_ftp_up_subword)
```

### Joint freq × FTP surface at final layer (bam)
```{r olmo-bam-joint, eval = params$source == "olmo"}
set.seed(964)
bam_joint_up_independently_final = cache_bam("bam_joint_up_independently_final", function() {
  bam(logit ~ te(log_freq, log_ftp) + s(verb_up, bs = 're'),
      data = up_indep_final_ftp, method = 'fREML', discrete = TRUE)
})
bam_joint_up_subword_final = cache_bam("bam_joint_up_subword_final", function() {
  bam(logit ~ te(log_freq, log_ftp) + s(verb_up, bs = 're'),
      data = up_sub_final_ftp, method = 'fREML', discrete = TRUE)
})
summary(bam_joint_up_independently_final)
summary(bam_joint_up_subword_final)
```

### Joint surface plots
```{r olmo-plot-joint, eval = params$source == "olmo"}
joint_indep   = plot_joint_surface(bam_joint_up_independently_final,
                                   up_indep_final_ftp, "Joint freq×FTP: UP independently (final layer)")
joint_subword = plot_joint_surface(bam_joint_up_subword_final,
                                   up_sub_final_ftp, "Joint freq×FTP: UP as subword (final layer)")
joint_indep$heatmap + joint_subword$heatmap
joint_indep$surface3d
joint_subword$surface3d
```

### Freq and FTP effects across layers (additive bam)
```{r olmo-bam-freq-ftp-layers, eval = params$source == "olmo"}
set.seed(964)
model_freq_ftp_layer_up_independently = cache_bam("model_freq_ftp_layer_up_independently", function() {
  bam(logit ~ te(log_freq, layer) + te(log_ftp, layer) + s(verb_up, bs = 're'),
      data = up_indep_ftp, method = 'fREML', discrete = TRUE)
})
model_freq_ftp_layer_up_subword = cache_bam("model_freq_ftp_layer_up_subword", function() {
  bam(logit ~ te(log_freq, layer) + te(log_ftp, layer) + s(verb_up, bs = 're'),
      data = up_subword_ftp, method = 'fREML', discrete = TRUE)
})
summary(model_freq_ftp_layer_up_independently)
summary(model_freq_ftp_layer_up_subword)
```

---

# BabyLM OPT Analysis

`r if (params$source != "babylm") "*(Not running — set params$source = 'babylm')*"`

## BabyLM plot helpers
```{r babylm-plot-helpers, eval = params$source == "babylm"}
plot_final_layer_by_model = function(model_fit, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)
  newdata = expand.grid(log_freq = log_freq_seq, model = levels(data$model)) %>%
    mutate(verb_up = NA)
  epred = add_epred_draws(newdata, model_fit, re_formula = NA, ndraws = 500)
  epred %>%
    ggplot(aes(x = log_freq, y = .epred, color = model, fill = model)) +
    stat_lineribbon(.width = 0.95, alpha = 0.2) +
    stat_summary(data = data, aes(x = log_freq, y = logit, color = model),
                 fun = mean, geom = "point", size = 1, alpha = 0.5) +
    labs(title = title, x = "log(frequency)", y = "Predicted logit",
         fill = "Model", color = "Model") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 11, face = "bold", margin = margin(b = 8)),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}

plot_freq_by_layer_by_model = function(model_fit, data, condition_label) {
  layer_norm_seq = seq(0, 1, length.out = 50)
  log_freq_seq   = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo    = quantile(data$log_freq, 0.1)
  log_freq_hi    = quantile(data$log_freq, 0.9)
  newdata_sp = expand.grid(layer_norm = layer_norm_seq, log_freq = log_freq_seq,
                            model = levels(data$model)) %>%
    mutate(verb_up = data$verb_up[1])
  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit
  p_sp = newdata_sp %>%
    ggplot(aes(x = log_freq, y = fit, color = layer_norm,
               group = interaction(model, layer_norm))) +
    geom_line(linewidth = 0.5, alpha = 0.7) + facet_wrap(~model) +
    scale_color_viridis_c(option = "plasma", name = "Layer\n(normalized)") +
    labs(title = paste0(condition_label, ": freq effect by layer (normalized) — by model"),
         x = "log(frequency)", y = "Predicted logit") + theme_minimal()
  newdata_diff = expand.grid(layer_norm = layer_norm_seq, model = levels(data$model),
                              freq_level = c("low", "high")) %>%
    mutate(log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
           verb_up  = data$verb_up[1])
  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit; newdata_diff$se = preds_diff$se.fit
  p_diff = newdata_diff %>%
    select(layer_norm, model, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(diff = fit_high - fit_low, se_diff = sqrt(se_high^2 + se_low^2),
           lo95 = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff) %>%
    ggplot(aes(x = layer_norm, y = diff, color = model, fill = model)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(condition_label, ": frequency effect (high − low) across layers — by model"),
         x = "Normalized layer (0 = first, 1 = final)",
         y = "Difference in predicted logit", color = "Model", fill = "Model") +
    theme_minimal()
  list(spaghetti = p_sp, diff = p_diff)
}

plot_ftp_by_layer_by_model = function(model_fit, data, condition_label) {
  layer_norm_seq = seq(0, 1, length.out = 50)
  log_ftp_seq    = seq(min(data$log_ftp, na.rm = TRUE), max(data$log_ftp, na.rm = TRUE), length.out = 50)
  ftp_lo = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi = quantile(data$log_ftp, 0.9, na.rm = TRUE)
  newdata_sp = expand.grid(layer_norm = layer_norm_seq, log_ftp = log_ftp_seq,
                            model = levels(data$model)) %>%
    mutate(verb_up = data$verb_up[1])
  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit
  p_sp = newdata_sp %>%
    ggplot(aes(x = log_ftp, y = fit, color = layer_norm,
               group = interaction(model, layer_norm))) +
    geom_line(linewidth = 0.5, alpha = 0.7) + facet_wrap(~model) +
    scale_color_viridis_c(option = "plasma", name = "Layer\n(normalized)") +
    labs(title = paste0(condition_label, ": FTP effect by layer — by model"),
         x = "log(FTP)", y = "Predicted logit") + theme_minimal()
  newdata_diff = expand.grid(layer_norm = layer_norm_seq, model = levels(data$model),
                              ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi), verb_up = data$verb_up[1])
  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit; newdata_diff$se = preds_diff$se.fit
  p_diff = newdata_diff %>%
    select(layer_norm, model, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(diff = fit_high - fit_low, se_diff = sqrt(se_high^2 + se_low^2),
           lo95 = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff) %>%
    ggplot(aes(x = layer_norm, y = diff, color = model, fill = model)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(condition_label, ": FTP effect (high − low) across layers — by model"),
         x = "Normalized layer (0 = first, 1 = final)",
         y = "Difference in predicted logit", color = "Model", fill = "Model") +
    theme_minimal()
  list(spaghetti = p_sp, diff = p_diff)
}
```

## UP independently

### Frequency at the final layer — by model (brms)
```{r babylm-brms-freq-indep-final, eval = params$source == "babylm"}
model_freq_up_independently = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_indep_final,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_independently")
)
summary(model_freq_up_independently)
```

### Frequency at the first layer — by model (brms)
```{r babylm-brms-freq-indep-first, eval = params$source == "babylm"}
model_freq_up_independently_first_layer = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_indep_first,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_independently_first_layer")
)
summary(model_freq_up_independently_first_layer)
```

### Frequency at the final layer — by model (bam — non-linear)
```{r babylm-bam-freq-indep-final, eval = params$source == "babylm"}
set.seed(964)
bam_freq_up_independently_final = cache_bam("bam_freq_up_independently_final", function() {
  bam(logit ~ model + s(log_freq, by = model) + s(verb_up, bs = 're'),
      data = up_indep_final, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_independently_final)
```

### Frequency at the first layer — by model (bam — non-linear)
```{r babylm-bam-freq-indep-first, eval = params$source == "babylm"}
set.seed(964)
bam_freq_up_independently_first = cache_bam("bam_freq_up_independently_first", function() {
  bam(logit ~ model + s(log_freq, by = model) + s(verb_up, bs = 're'),
      data = up_indep_first, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_independently_first)
```

### Frequency × layer — by model (mgcv, normalized layers)
```{r babylm-bam-freq-layer-indep, eval = params$source == "babylm"}
set.seed(964)
model_freq_layer_up_independently = cache_bam("model_freq_layer_up_independently", function() {
  bam(logit ~ model + te(log_freq, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_independently, method = 'fREML', discrete = TRUE)
})
summary(model_freq_layer_up_independently)
```

## UP as a subword

### Frequency at the final layer — by model (brms)
```{r babylm-brms-freq-sub-final, eval = params$source == "babylm"}
model_freq_up_subword = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_sub_final,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_subword")
)
summary(model_freq_up_subword)
```

### Frequency at the first layer — by model (brms)
```{r babylm-brms-freq-sub-first, eval = params$source == "babylm"}
model_freq_up_subword_first_layer = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_sub_first,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_up_subword_first_layer")
)
summary(model_freq_up_subword_first_layer)
```

### Frequency at the final layer — by model (bam — non-linear)
```{r babylm-bam-freq-sub-final, eval = params$source == "babylm"}
set.seed(964)
bam_freq_up_subword_final = cache_bam("bam_freq_up_subword_final", function() {
  bam(logit ~ model + s(log_freq, k = 10, by = model) + s(verb_up, bs = 're'),
      data = up_sub_final, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_subword_final)
```

### Frequency at the first layer — by model (bam — non-linear)
```{r babylm-bam-freq-sub-first, eval = params$source == "babylm"}
set.seed(964)
bam_freq_up_subword_first = cache_bam("bam_freq_up_subword_first", function() {
  bam(logit ~ model + s(log_freq, k = 10, by = model) + s(verb_up, bs = 're'),
      data = up_sub_first, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_up_subword_first)
```

### Frequency × layer — by model (mgcv, normalized layers)
```{r babylm-bam-freq-layer-sub, eval = params$source == "babylm"}
set.seed(964)
model_freq_layer_up_subword = cache_bam("model_freq_layer_up_subword", function() {
  bam(logit ~ model + te(log_freq, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_subword, method = 'fREML', discrete = TRUE)
})
summary(model_freq_layer_up_subword)
```

## BabyLM frequency plots
```{r babylm-freq-plots, eval = params$source == "babylm"}
p_indep_final = plot_final_layer_by_model(model_freq_up_independently,
  up_indep_final, "UP independently: frequency effect at final layer — by model")
p_sub_final   = plot_final_layer_by_model(model_freq_up_subword,
  up_sub_final,   "UP as subword: frequency effect at final layer — by model")
p_indep_first = plot_final_layer_by_model(model_freq_up_independently_first_layer,
  up_indep_first, "UP independently: frequency effect at first layer — by model")
p_sub_first   = plot_final_layer_by_model(model_freq_up_subword_first_layer,
  up_sub_first,   "UP as subword: frequency effect at first layer — by model")
(p_indep_first | p_sub_first) / (p_indep_final | p_sub_final)

plots_indep   = plot_freq_by_layer_by_model(model_freq_layer_up_independently,
                                             up_independently, "UP independently")
plots_subword = plot_freq_by_layer_by_model(model_freq_layer_up_subword,
                                             up_subword, "UP as subword")
plots_indep$spaghetti;   plots_indep$diff
plots_subword$spaghetti; plots_subword$diff
```

---

## BabyLM — Predictability (FTP = P(up|V))

### Effect of FTP at the final layer — by model (brms)
```{r babylm-brms-ftp-final, eval = params$source == "babylm"}
up_indep_final_ftp = up_indep_final %>% filter(!is.na(log_ftp))
up_sub_final_ftp   = up_sub_final   %>% filter(!is.na(log_ftp))
up_indep_first_ftp = up_indep_first %>% filter(!is.na(log_ftp))
up_sub_first_ftp   = up_sub_first   %>% filter(!is.na(log_ftp))

model_ftp_up_independently = brm(
  formula = logit ~ log_ftp * model + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_ftp_up_independently")
)
model_ftp_up_independently_first_layer = brm(
  formula = logit ~ log_ftp * model + (1 | verb_up),
  data    = up_indep_first_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_ftp_up_independently_first_layer")
)
summary(model_ftp_up_independently)
summary(model_ftp_up_independently_first_layer)
```

### Effect of FTP at the final layer — by model (bam — non-linear)
```{r babylm-bam-ftp-final, eval = params$source == "babylm"}
set.seed(964)
bam_ftp_up_independently_final = cache_bam("bam_ftp_up_independently_final", function() {
  bam(logit ~ model + s(log_ftp, by = model) + s(verb_up, bs = 're'),
      data = up_indep_final_ftp, method = 'fREML', discrete = TRUE)
})
bam_ftp_up_subword_final = cache_bam("bam_ftp_up_subword_final", function() {
  bam(logit ~ model + s(log_ftp, by = model) + s(verb_up, bs = 're'),
      data = up_sub_final_ftp, method = 'fREML', discrete = TRUE)
})
summary(bam_ftp_up_independently_final)
summary(bam_ftp_up_subword_final)
```

### Effect of FTP across layers — by model (mgcv, normalized layers)
```{r babylm-bam-ftp-layers, eval = params$source == "babylm"}
up_indep_ftp   = up_independently %>% filter(!is.na(log_ftp))
up_subword_ftp = up_subword       %>% filter(!is.na(log_ftp))

set.seed(964)
model_ftp_layer_up_independently = cache_bam("model_ftp_layer_up_independently", function() {
  bam(logit ~ model + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_indep_ftp, method = 'fREML', discrete = TRUE)
})
model_ftp_layer_up_subword = cache_bam("model_ftp_layer_up_subword", function() {
  bam(logit ~ model + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_subword_ftp, method = 'fREML', discrete = TRUE)
})
summary(model_ftp_layer_up_independently)
summary(model_ftp_layer_up_subword)
```

### FTP plots — by model
```{r babylm-ftp-plots, eval = params$source == "babylm"}
plots_ftp_indep   = plot_ftp_by_layer_by_model(model_ftp_layer_up_independently,
                                                up_indep_ftp, "UP independently")
plots_ftp_subword = plot_ftp_by_layer_by_model(model_ftp_layer_up_subword,
                                                up_subword_ftp, "UP as subword")
plots_ftp_indep$spaghetti;   plots_ftp_indep$diff
plots_ftp_subword$spaghetti; plots_ftp_subword$diff
```

---

## BabyLM — Joint frequency + predictability

### Additive freq + FTP at the final layer — by model (brms)
```{r babylm-brms-joint, eval = params$source == "babylm"}
model_freq_ftp_up_independently = brm(
  formula = logit ~ log_freq * model + log_ftp * model + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_ftp_up_independently")
)
model_freq_ftp_up_subword = brm(
  formula = logit ~ log_freq * model + log_ftp * model + (1 | verb_up),
  data    = up_sub_final_ftp,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_ftp_up_subword")
)
summary(model_freq_ftp_up_independently)
summary(model_freq_ftp_up_subword)
```

### Joint freq × FTP surface at final layer — by model (bam)
```{r babylm-bam-joint, eval = params$source == "babylm"}
set.seed(964)
bam_joint_up_independently_final = cache_bam("bam_joint_up_independently_final", function() {
  bam(logit ~ model + te(log_freq, log_ftp, by = model) + s(verb_up, bs = 're'),
      data = up_indep_final_ftp, method = 'fREML', discrete = TRUE)
})
bam_joint_up_subword_final = cache_bam("bam_joint_up_subword_final", function() {
  bam(logit ~ model + te(log_freq, log_ftp, by = model) + s(verb_up, bs = 're'),
      data = up_sub_final_ftp, method = 'fREML', discrete = TRUE)
})
summary(bam_joint_up_independently_final)
summary(bam_joint_up_subword_final)
```

### Freq and FTP across layers — by model (additive bam)
```{r babylm-bam-freq-ftp-layers, eval = params$source == "babylm"}
set.seed(964)
model_freq_ftp_layer_up_independently = cache_bam("model_freq_ftp_layer_up_independently", function() {
  bam(logit ~ model + te(log_freq, layer_norm, by = model) +
        te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_indep_ftp, method = 'fREML', discrete = TRUE)
})
model_freq_ftp_layer_up_subword = cache_bam("model_freq_ftp_layer_up_subword", function() {
  bam(logit ~ model + te(log_freq, layer_norm, by = model) +
        te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
      data = up_subword_ftp, method = 'fREML', discrete = TRUE)
})
summary(model_freq_ftp_layer_up_independently)
summary(model_freq_ftp_layer_up_subword)
```

---

# Whisper-small Analysis

`r if (params$source != "whisper") "*(Not running — set params$source = 'whisper')*"`

Whisper embeddings are extracted via a single forward pass per utterance:

- **Encoder**: the hidden states at a given layer are mean-pooled over the audio
  frames spanning the "up" word's time window (WhisperX word-level timestamps,
  20 ms per encoder output frame). The resulting vector is the "up" embedding.
- **Decoder**: hidden states are extracted at the exact token position of "up"
  using teacher-forced decoding (the gold transcript is fed as decoder input).

A separate logistic regression classifier is trained **at each layer
independently** on standalone "up" (positive) vs. random other-word (negative)
embeddings. The classifier trained at layer *L* is then applied to V+up test
embeddings extracted from that same layer *L*. No classifier trained on one
layer is ever applied to another layer's embeddings.

## Effect of frequency at the final layer — by component (brms)
```{r whisper-brms-freq-final, eval = params$source == "whisper"}
model_freq_final = brm(
  formula = logit ~ log_freq * component + (1 | verb_up),
  data    = whisper_final,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_final")
)
summary(model_freq_final)
```

## Effect of frequency at the first layer — by component (brms)
```{r whisper-brms-freq-first, eval = params$source == "whisper"}
model_freq_first = brm(
  formula = logit ~ log_freq * component + (1 | verb_up),
  data    = whisper_first,
  prior   = priors,
  iter = 6000, warmup = 3000, chains = 4, cores = 4, seed = 964,
  file    = brms_path("model_freq_first")
)
summary(model_freq_first)
```

## Effect of frequency at the final layer — by component (bam — non-linear)
```{r whisper-bam-freq-final, eval = params$source == "whisper"}
set.seed(964)
bam_freq_final = cache_bam("bam_freq_final", function() {
  bam(logit ~ component + s(log_freq, by = component) + s(verb_up, bs = 're'),
      data = whisper_final, method = 'fREML', discrete = TRUE)
})
summary(bam_freq_final)
```

## Effect of frequency across all layers — by component (mgcv)
```{r whisper-bam-freq-layer, eval = params$source == "whisper"}
set.seed(964)
model_freq_layer = cache_bam("model_freq_layer", function() {
  bam(logit ~ component + te(log_freq, layer, by = component) + s(verb_up, bs = 're'),
      data = whisper_all, method = 'fREML', discrete = TRUE)
})
summary(model_freq_layer)
```

## Whisper plots

### Final and first layer — frequency effect by component (brms)
```{r whisper-plot-brms-freq, eval = params$source == "whisper"}
plot_layer_by_component = function(model_fit, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)
  newdata = expand.grid(log_freq = log_freq_seq, component = levels(data$component)) %>%
    mutate(verb_up = NA)
  epred = add_epred_draws(newdata, model_fit, re_formula = NA, ndraws = 500)
  epred %>%
    ggplot(aes(x = log_freq, y = .epred, color = component, fill = component)) +
    stat_lineribbon(.width = 0.95, alpha = 0.2) +
    stat_summary(data = data, aes(x = log_freq, y = logit, color = component),
                 fun = mean, geom = "point", size = 1, alpha = 0.5) +
    labs(title = title, x = "log(frequency)", y = "Predicted logit",
         fill = "Component", color = "Component") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(size = 11, face = "bold", margin = margin(b = 8)),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}

p_final = plot_layer_by_component(model_freq_final, whisper_final,
  "Whisper-small: frequency effect at final layer (11) — encoder vs decoder")
p_first = plot_layer_by_component(model_freq_first, whisper_first,
  "Whisper-small: frequency effect at first layer (0) — encoder vs decoder")
p_first / p_final
```

### Across layers — frequency effect by component (mgcv)
```{r whisper-plot-freq-layers, eval = params$source == "whisper"}
plot_freq_by_layer_whisper = function(model_fit, data, title) {
  layer_seq    = 0:11
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo  = quantile(data$log_freq, 0.1)
  log_freq_hi  = quantile(data$log_freq, 0.9)

  newdata_sp = expand.grid(layer = layer_seq, log_freq = log_freq_seq,
                            component = levels(data$component)) %>%
    mutate(verb_up = data$verb_up[1])
  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit
  p_sp = newdata_sp %>%
    ggplot(aes(x = log_freq, y = fit, color = layer,
               group = interaction(component, layer))) +
    geom_line(linewidth = 0.5, alpha = 0.7) + facet_wrap(~component) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(title = paste0(title, ": frequency effect by layer — encoder vs decoder"),
         x = "log(frequency)", y = "Predicted logit") + theme_minimal()

  newdata_diff = expand.grid(layer = layer_seq, component = levels(data$component),
                              freq_level = c("low", "high")) %>%
    mutate(log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
           verb_up  = data$verb_up[1])
  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit; newdata_diff$se = preds_diff$se.fit
  p_diff = newdata_diff %>%
    select(layer, component, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(diff = fit_high - fit_low, se_diff = sqrt(se_high^2 + se_low^2),
           lo95 = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff) %>%
    ggplot(aes(x = layer, y = diff, color = component, fill = component)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(title, ": frequency effect (high − low) across layers"),
         x = "Layer", y = "Difference in predicted logit (90th − 10th pct freq)",
         color = "Component", fill = "Component") + theme_minimal()

  newdata_heat = expand.grid(layer = layer_seq, log_freq = log_freq_seq,
                              component = levels(data$component)) %>%
    mutate(verb_up = data$verb_up[1])
  preds_heat = predict(model_fit, newdata = newdata_heat, se.fit = FALSE, exclude = 's(verb_up)')
  newdata_heat$fit = preds_heat
  p_heat = newdata_heat %>%
    ggplot(aes(x = layer, y = log_freq, fill = fit)) + geom_tile() +
    facet_wrap(~component) +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = paste0(title, ": predicted logit surface"), x = "Layer", y = "log(frequency)") +
    theme_minimal()

  list(spaghetti = p_sp, diff = p_diff, heatmap = p_heat)
}

plots_whisper = plot_freq_by_layer_whisper(model_freq_layer, whisper_all, "Whisper-small")
plots_whisper$spaghetti
plots_whisper$diff
plots_whisper$heatmap
```

### Classifier validation accuracy across layers — encoder vs decoder
```{r whisper-val-accuracy, eval = params$source == "whisper"}
read_layer_meta = function(path, component) {
  meta = jsonlite::fromJSON(path)
  tibble(
    layer       = sapply(meta$layers, `[[`, "layer"),
    component   = component,
    val_acc     = sapply(meta$layers, `[[`, "val_acc"),
    val_up_acc  = sapply(meta$layers, `[[`, "val_up_acc"),
    val_oth_acc = sapply(meta$layers, `[[`, "val_other_acc"),
    cv_mean     = sapply(meta$layers, `[[`, "cv_mean"),
  )
}

meta_enc = read_layer_meta('../Data/whisper/encoder/layer_metadata.json', 'encoder')
meta_dec = read_layer_meta('../Data/whisper/decoder/layer_metadata.json', 'decoder')
meta = bind_rows(meta_enc, meta_dec) %>%
  mutate(component = factor(component, levels = c("encoder", "decoder")))

meta %>%
  ggplot(aes(x = layer, y = val_acc, color = component)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_color_manual(values = c("encoder" = "#1a9850", "decoder" = "#d73027")) +
  labs(title = "Whisper-small: classifier validation accuracy across layers",
       x = "Layer", y = "Validation accuracy", color = "Component") +
  theme_minimal(base_size = 13)
```

### Spearman correlation between log(frequency) and mean logit — across layers
```{r whisper-spearman, eval = params$source == "whisper"}
summary_by_layer = whisper_all %>%
  group_by(component, layer, verb_up, frequency) %>%
  summarise(mean_logit = mean(logit), .groups = "drop")

cor_by_layer = summary_by_layer %>%
  group_by(component, layer) %>%
  summarise(spearman_r = cor(log(frequency), mean_logit, method = "spearman"),
            .groups = "drop")

cor_by_layer %>%
  ggplot(aes(x = layer, y = spearman_r, color = component)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("encoder" = "#1a9850", "decoder" = "#d73027")) +
  labs(title = "Spearman r(log freq, mean logit) across layers — encoder vs decoder",
       subtitle = "Negative: high-freq V+up types look less like standalone 'up'",
       x = "Layer", y = "Spearman r", color = "Component") +
  theme_minimal(base_size = 13)
```

### Most and least compositional V+up types — final layer
```{r whisper-top-bottom, eval = params$source == "whisper"}
plot_top_bottom = function(data, component_name, layer_num = 11) {
  d = data %>%
    filter(component == component_name, layer == layer_num) %>%
    group_by(verb_up, frequency) %>%
    summarise(mean_logit = mean(logit), .groups = "drop")
  n_show = 20
  bar_df = bind_rows(
    d %>% slice_max(mean_logit, n = n_show),
    d %>% slice_min(mean_logit, n = n_show)
  ) %>% distinct(verb_up, .keep_all = TRUE) %>% arrange(mean_logit)
  colors = ifelse(bar_df$mean_logit < 0, "#d73027", "#1a9850")
  bar_df %>%
    ggplot(aes(x = mean_logit, y = reorder(verb_up, mean_logit))) +
    geom_col(fill = colors, alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(title = paste0(component_name, " (layer ", layer_num,
                        "): most & least compositional V+up"),
         x = "Mean logit (higher = more standalone-'up'-like)", y = NULL) +
    theme_minimal(base_size = 11)
}

plot_top_bottom(whisper_all, "encoder") /
plot_top_bottom(whisper_all, "decoder")
```
