---
title: "analysis-script"
author: "Zachary Houghton"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
```

# analysis

## load data

```{r}
up_independently = read_csv('../Data/Data_up/all_layers_results.csv')
up_subword = read_csv('../Data/Data_upsubword/all_layers_results.csv')

up_independently = up_independently %>%
  mutate(
    log_freq = log(frequency),
    verb_up  = factor(verb_up)
  )

up_subword = up_subword %>%
  mutate(
    log_freq = log(frequency),
    verb_up  = factor(verb_up)
  )
```

## up independently

### Effect of frequency at the final hidden layer (brms)

```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_up_independently = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_independently %>% filter(layer == 31),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_independently",
  seed = 964
)
```

### Effect of frequency at the first hidden layer (brms)

```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_up_independently_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_independently %>% filter(layer == 0),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_independently_first_layer",
  seed = 964
)
```

### Effect of frequency interacting with layer (mgcv)

```{r}
# bam() is the faster version of gam() for large datasets
# verb_up as a random effect via s(verb_up, bs = 're')
set.seed(964)
if (file.exists("model_freq_layer_up_independently.rds")) {
  model_freq_layer_up_independently = readRDS("model_freq_layer_up_independently.rds")
} else {
  model_freq_layer_up_independently = bam(
    formula  = logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
    data     = up_independently,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_independently, "model_freq_layer_up_independently.rds")
}

summary(model_freq_layer_up_independently)
```

## up as a subword

### Effect of frequency at the final hidden layer (brms)

```{r}
model_freq_up_subword = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_subword %>% filter(layer == 31),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_subword",
  seed = 964
)
```

### Effect of frequency at the first hidden layer (brms)

```{r}
model_freq_up_subword_first_layer = brm(
  formula = logit ~ log_freq + (1 | verb_up),
  data = up_subword %>% filter(layer == 0),
  prior = priors,
  iter = 6000,
  warmup = 3000,
  chains = 4,
  cores = 4,
  file = "model_freq_up_subword_first_layer",
  seed = 964
)
```

### Effect of frequency interacting with layer (mgcv)

```{r}
set.seed(964)
if (file.exists("model_freq_layer_up_subword.rds")) {
  model_freq_layer_up_subword = readRDS("model_freq_layer_up_subword.rds")
} else {
  model_freq_layer_up_subword = bam(
    formula  = logit ~ te(log_freq, layer) + s(verb_up, bs = 're'),
    data     = up_subword,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_subword, "model_freq_layer_up_subword.rds")
}

summary(model_freq_layer_up_subword)
```

## plots

### final and first layer plots

```{r}
library(tidybayes)
library(patchwork)
library(viridis)
library(plotly)

plot_final_layer = function(model, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)
  
  newdata = data.frame(
    log_freq = log_freq_seq,
    verb_up  = NA
  )
  
  epred = add_epred_draws(newdata, model, re_formula = NA, ndraws = 500)
  
  epred %>%
    ggplot(aes(x = log_freq, y = .epred)) +
    stat_lineribbon(.width = c(0.95, 0.80, 0.50), alpha = 0.35) +
    scale_fill_manual(
      values = c("0.95" = "#b3cde3", "0.80" = "#6497b1", "0.50" = "#005b96"),
      labels = c("95%", "80%", "50%")
    ) +
    stat_summary(
      data = data, aes(x = log_freq, y = logit),
      fun      = mean,
      geom     = "point",
      color    = "gray30",
      size     = 1.5,
      alpha    = 0.6
    ) +
    labs(
      title = title,
      x     = "log(frequency)",
      y     = "Predicted logit",
      fill  = "Credible interval"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold", margin = margin(b = 8)),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom",
      legend.key.size  = unit(0.5, "cm")
    )
}

# Final layer
p_indep_final = plot_final_layer(
  model_freq_up_independently,
  up_independently %>% filter(layer == 31),
  "UP independently: frequency effect at final layer (31)"
)

p_subword_final = plot_final_layer(
  model_freq_up_subword,
  up_subword %>% filter(layer == 31),
  "UP as subword: frequency effect at final layer (31)"
)

# First layer
p_indep_first = plot_final_layer(
  model_freq_up_independently_first_layer,
  up_independently %>% filter(layer == 0),
  "UP independently: frequency effect at first layer (0)"
)

p_subword_first = plot_final_layer(
  model_freq_up_subword_first_layer,
  up_subword %>% filter(layer == 0),
  "UP as subword: frequency effect at first layer (0)"
)

# Arrange: rows = layer (first vs final), columns = condition (indep vs subword)
(p_indep_first  | p_subword_first) /
(p_indep_final  | p_subword_final)
```

### across layers

```{r}
plot_freq_by_layer_mgcv = function(model, data, title) {
  
  layer_seq    = 0:31
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo  = quantile(data$log_freq, 0.1)
  log_freq_hi  = quantile(data$log_freq, 0.9)
  
  # ── 1. Spaghetti: one line per layer across log_freq range ────────────────
  newdata_spaghetti = expand.grid(
    layer    = layer_seq,
    log_freq = log_freq_seq
  ) %>%
    mutate(verb_up = data$verb_up[1])
  
  preds_spaghetti = predict(
    model,
    newdata = newdata_spaghetti,
    se.fit  = TRUE,
    exclude = 's(verb_up)'
  )
  
  newdata_spaghetti = newdata_spaghetti %>%
    mutate(
      fit  = preds_spaghetti$fit,
      se   = preds_spaghetti$se.fit,
      lo95 = fit - 1.96 * se,
      hi95 = fit + 1.96 * se
    )
  
  p_spaghetti = newdata_spaghetti %>%
    ggplot(aes(x = log_freq, y = fit, color = layer, group = factor(layer))) +
    geom_line(linewidth = 0.7, alpha = 0.8) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(
      title = paste0(title, ": effect of log(freq) by layer"),
      x     = "log(frequency)",
      y     = "Predicted logit"
    ) +
    theme_minimal()
  
  # ── 2. Heatmap: logit surface across freq x layer ─────────────────────────
  p_heatmap = newdata_spaghetti %>%
    ggplot(aes(x = layer, y = log_freq, fill = fit)) +
    geom_tile() +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(
      title = paste0(title, ": predicted logit surface"),
      x     = "Layer",
      y     = "log(frequency)"
    ) +
    theme_minimal()
  
  # ── 3. Difference plot: frequency effect (high - low) across layers ────────
  newdata_diff = expand.grid(
    layer      = layer_seq,
    freq_level = c("low", "high")
  ) %>%
    mutate(
      log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
      verb_up  = data$verb_up[1]
    )
  
  preds_diff = predict(
    model,
    newdata = newdata_diff,
    se.fit  = TRUE,
    exclude = 's(verb_up)'
  )
  
  newdata_diff = newdata_diff %>%
    mutate(fit = preds_diff$fit, se = preds_diff$se.fit)
  
  p_diff = newdata_diff %>%
    select(layer, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff,
      lo80    = diff - 1.28 * se_diff,
      hi80    = diff + 1.28 * se_diff
    ) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(title, ": frequency effect (high - low) across layers"),
      x     = "Layer",
      y     = "Difference in predicted logit (90th - 10th pct freq)"
    ) +
    theme_minimal()
  
  # ── 4. 3D surface (plotly) ─────────────────────────────────────────────────
  fit_matrix = newdata_spaghetti %>%
    select(layer, log_freq, fit) %>%
    pivot_wider(names_from = layer, values_from = fit) %>%
    arrange(log_freq) %>%
    select(-log_freq) %>%
    as.matrix()
  
  p_3d = plot_ly(
    x = ~layer_seq,
    y = ~log_freq_seq,
    z = ~fit_matrix,
    type       = "surface",
    colorscale = "Viridis",
    showscale  = TRUE
  ) %>%
    layout(
      title = paste0(title, ": 3D predicted logit surface"),
      scene = list(
        xaxis = list(title = "Layer"),
        yaxis = list(title = "log(frequency)"),
        zaxis = list(title = "Predicted logit")
      )
    )
  
  list(
    spaghetti = p_spaghetti,
    heatmap   = p_heatmap,
    diff      = p_diff,
    surface3d = p_3d
  )
}

plots_indep = plot_freq_by_layer_mgcv(
  model_freq_layer_up_independently,
  up_independently,
  "UP independently"
)

plots_subword = plot_freq_by_layer_mgcv(
  model_freq_layer_up_subword,
  up_subword,
  "UP as subword"
)

# 2D plots combined
(plots_indep$spaghetti + plots_subword$spaghetti) /
(plots_indep$heatmap   + plots_subword$heatmap)   /
(plots_indep$diff      + plots_subword$diff)

# 3D plots separately (plotly doesn't compose with patchwork)
plots_indep$surface3d
plots_subword$surface3d
```