---
title: "analysis-script — BabyLM OPT models (125m / 350m / 1.3b)"
author: "Zachary Houghton"
date: "2026-03-03"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)
library(plotly)
```

# Analysis — BabyLM OPT Models

## Load data

Each model's results are loaded from its own subdirectory. A `model` column is
added before combining so all three can be compared in a single analysis.

OPT-125m has 12 layers (0–11); OPT-350m and OPT-1.3b each have 24 layers (0–23).

```{r}
FINAL_LAYER = list(
  "opt-125m" = 11,
  "opt-350m" = 23,
  "opt-1.3b" = 23
)

load_model_data = function(model_name, subdir) {
  up_indep = read_csv(paste0('../../Data/babylm/', subdir, '/Data_up/all_layers_results.csv')) %>%
    mutate(model = model_name, log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))
  up_sub = read_csv(paste0('../../Data/babylm/', subdir, '/Data_upsubword/all_layers_results.csv')) %>%
    mutate(model = model_name, log_freq = log(frequency), log_ftp = log(ftp), verb_up = factor(verb_up))
  list(up_indep = up_indep, up_sub = up_sub)
}

d_125m = load_model_data("opt-125m", "opt-125m")
d_350m = load_model_data("opt-350m", "opt-350m")
d_1.3b = load_model_data("opt-1.3b", "opt-1.3b")

up_independently = bind_rows(d_125m$up_indep, d_350m$up_indep, d_1.3b$up_indep) %>%
  mutate(model = factor(model, levels = c("opt-125m", "opt-350m", "opt-1.3b")))

up_subword = bind_rows(d_125m$up_sub, d_350m$up_sub, d_1.3b$up_sub) %>%
  mutate(model = factor(model, levels = c("opt-125m", "opt-350m", "opt-1.3b")))

# Per-model final-layer subsets (used in single-layer analyses)
up_indep_final = up_independently %>%
  group_by(model) %>%
  filter(layer == FINAL_LAYER[[as.character(model[1])]]) %>%
  ungroup()

up_sub_final = up_subword %>%
  group_by(model) %>%
  filter(layer == FINAL_LAYER[[as.character(model[1])]]) %>%
  ungroup()
```

## UP independently

### Effect of frequency at the final hidden layer — by model (brms)
```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_up_independently = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_indep_final,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_up_independently",
  seed    = 964
)
```

### Effect of frequency at the first hidden layer — by model (brms)
```{r}
up_indep_first = up_independently %>% filter(layer == 0)

model_freq_up_independently_first_layer = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_indep_first,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_up_independently_first_layer",
  seed    = 964
)
```

### Effect of frequency at the final hidden layer — by model (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_independently_final.rds")) {
  bam_freq_up_independently_final = readRDS("bam_freq_up_independently_final.rds")
} else {
  bam_freq_up_independently_final = bam(
    formula  = logit ~ model + s(log_freq, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_final,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_independently_final, "bam_freq_up_independently_final.rds")
}

summary(bam_freq_up_independently_final)
```

### Effect of frequency at the first hidden layer — by model (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_independently_first.rds")) {
  bam_freq_up_independently_first = readRDS("bam_freq_up_independently_first.rds")
} else {
  bam_freq_up_independently_first = bam(
    formula  = logit ~ model + s(log_freq, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_first,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_independently_first, "bam_freq_up_independently_first.rds")
}

summary(bam_freq_up_independently_first)
```

### Effect of frequency interacting with layer — by model (mgcv)

Note: layers are on different scales across models (125m: 0–11, 350m/1.3b: 0–23).
We use `layer_norm` (layer / max_layer) so all models share a [0, 1] scale.

```{r}
up_independently = up_independently %>%
  group_by(model) %>%
  mutate(layer_norm = layer / max(layer)) %>%
  ungroup()

set.seed(964)
if (file.exists("model_freq_layer_up_independently.rds")) {
  model_freq_layer_up_independently = readRDS("model_freq_layer_up_independently.rds")
} else {
  model_freq_layer_up_independently = bam(
    formula  = logit ~ model + te(log_freq, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_independently,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_independently, "model_freq_layer_up_independently.rds")
}

summary(model_freq_layer_up_independently)
```

## UP as a subword

### Effect of frequency at the final hidden layer — by model (brms)
```{r}
model_freq_up_subword = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_sub_final,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_up_subword",
  seed    = 964
)
```

### Effect of frequency at the first hidden layer — by model (brms)
```{r}
up_sub_first = up_subword %>% filter(layer == 0)

model_freq_up_subword_first_layer = brm(
  formula = logit ~ log_freq * model + (1 | verb_up),
  data    = up_sub_first,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_up_subword_first_layer",
  seed    = 964
)
```

### Effect of frequency at the final hidden layer — by model (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_subword_final.rds")) {
  bam_freq_up_subword_final = readRDS("bam_freq_up_subword_final.rds")
} else {
  bam_freq_up_subword_final = bam(
    formula  = logit ~ model + s(log_freq, k = 10, by = model) + s(verb_up, bs = 're'),
    data     = up_sub_final,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_subword_final, "bam_freq_up_subword_final.rds")
}

summary(bam_freq_up_subword_final)
```

### Effect of frequency at the first hidden layer — by model (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_up_subword_first.rds")) {
  bam_freq_up_subword_first = readRDS("bam_freq_up_subword_first.rds")
} else {
  bam_freq_up_subword_first = bam(
    formula  = logit ~ model + s(log_freq, k = 10, by = model) + s(verb_up, bs = 're'),
    data     = up_sub_first,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_up_subword_first, "bam_freq_up_subword_first.rds")
}

summary(bam_freq_up_subword_first)
```

### Effect of frequency interacting with layer — by model (mgcv)
```{r}
up_subword = up_subword %>%
  group_by(model) %>%
  mutate(layer_norm = layer / max(layer)) %>%
  ungroup()

set.seed(964)
if (file.exists("model_freq_layer_up_subword.rds")) {
  model_freq_layer_up_subword = readRDS("model_freq_layer_up_subword.rds")
} else {
  model_freq_layer_up_subword = bam(
    formula  = logit ~ model + te(log_freq, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_subword,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer_up_subword, "model_freq_layer_up_subword.rds")
}

summary(model_freq_layer_up_subword)
```

## Plots

### Final and first layer — frequency effect by model (brms)
```{r}
plot_final_layer_by_model = function(model_fit, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)

  newdata = expand.grid(
    log_freq = log_freq_seq,
    model    = levels(data$model)
  ) %>% mutate(verb_up = NA)

  epred = add_epred_draws(newdata, model_fit, re_formula = NA, ndraws = 500)

  epred %>%
    ggplot(aes(x = log_freq, y = .epred, color = model, fill = model)) +
    stat_lineribbon(.width = 0.95, alpha = 0.2) +
    stat_summary(
      data  = data,
      aes(x = log_freq, y = logit, color = model),
      fun   = mean,
      geom  = "point",
      size  = 1,
      alpha = 0.5
    ) +
    labs(
      title = title,
      x     = "log(frequency)",
      y     = "Predicted logit",
      fill  = "Model",
      color = "Model"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold", margin = margin(b = 8)),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}

p_indep_final = plot_final_layer_by_model(
  model_freq_up_independently,
  up_indep_final,
  "UP independently: frequency effect at final layer — by model"
)

p_subword_final = plot_final_layer_by_model(
  model_freq_up_subword,
  up_sub_final,
  "UP as subword: frequency effect at final layer — by model"
)

p_indep_first = plot_final_layer_by_model(
  model_freq_up_independently_first_layer,
  up_indep_first,
  "UP independently: frequency effect at first layer — by model"
)

p_subword_first = plot_final_layer_by_model(
  model_freq_up_subword_first_layer,
  up_sub_first,
  "UP as subword: frequency effect at first layer — by model"
)

(p_indep_first  | p_subword_first) /
(p_indep_final  | p_subword_final)
```

### Across layers — frequency effect by model (mgcv, normalized layers)
```{r}
plot_freq_by_layer_by_model = function(model_fit, data, condition_label) {
  layer_norm_seq = seq(0, 1, length.out = 50)
  log_freq_seq   = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo    = quantile(data$log_freq, 0.1)
  log_freq_hi    = quantile(data$log_freq, 0.9)

  # ── Spaghetti ──────────────────────────────────────────────────────────────
  newdata_sp = expand.grid(
    layer_norm = layer_norm_seq,
    log_freq   = log_freq_seq,
    model      = levels(data$model)
  ) %>% mutate(verb_up = data$verb_up[1])

  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE,
                     exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit

  p_sp = newdata_sp %>%
    ggplot(aes(x = log_freq, y = fit, color = layer_norm, group = interaction(model, layer_norm))) +
    geom_line(linewidth = 0.5, alpha = 0.7) +
    facet_wrap(~model) +
    scale_color_viridis_c(option = "plasma", name = "Layer\n(normalized)") +
    labs(
      title = paste0(condition_label, ": freq effect by layer (normalized) — by model"),
      x     = "log(frequency)",
      y     = "Predicted logit"
    ) +
    theme_minimal()

  # ── Difference (high - low freq) across normalized layers ─────────────────
  newdata_diff = expand.grid(
    layer_norm = layer_norm_seq,
    model      = levels(data$model),
    freq_level = c("low", "high")
  ) %>%
    mutate(
      log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
      verb_up  = data$verb_up[1]
    )

  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE,
                       exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit
  newdata_diff$se  = preds_diff$se.fit

  p_diff = newdata_diff %>%
    select(layer_norm, model, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff
    ) %>%
    ggplot(aes(x = layer_norm, y = diff, color = model, fill = model)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(condition_label, ": frequency effect (high - low) across layers — by model"),
      x     = "Normalized layer (0 = first, 1 = final)",
      y     = "Difference in predicted logit",
      color = "Model", fill = "Model"
    ) +
    theme_minimal()

  list(spaghetti = p_sp, diff = p_diff)
}

plots_indep   = plot_freq_by_layer_by_model(model_freq_layer_up_independently, up_independently,   "UP independently")
plots_subword = plot_freq_by_layer_by_model(model_freq_layer_up_subword,       up_subword, "UP as subword")

plots_indep$spaghetti
plots_indep$diff

plots_subword$spaghetti
plots_subword$diff
```

---

## Predictability (FTP = P(up|V))

FTP analyses require `--corpus-stats-pkl` (BabyLM corpus stats) to have been passed to the pipeline.
Rows with missing FTP are dropped for FTP-specific models.

### Effect of FTP at the final hidden layer — by model (brms)
```{r}
up_indep_final_ftp  = up_indep_final  %>% filter(!is.na(log_ftp))
up_sub_final_ftp    = up_sub_final    %>% filter(!is.na(log_ftp))
up_indep_first_ftp  = up_independently %>% filter(layer == 0, !is.na(log_ftp))
up_sub_first_ftp    = up_subword       %>% filter(layer == 0, !is.na(log_ftp))

model_ftp_up_independently = brm(
  formula = logit ~ log_ftp * model + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_ftp_up_independently",
  seed    = 964
)
```

### Effect of FTP at the first hidden layer — by model (brms)
```{r}
model_ftp_up_independently_first_layer = brm(
  formula = logit ~ log_ftp * model + (1 | verb_up),
  data    = up_indep_first_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_ftp_up_independently_first_layer",
  seed    = 964
)
```

### Effect of FTP at the final hidden layer — by model (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_ftp_up_independently_final.rds")) {
  bam_ftp_up_independently_final = readRDS("bam_ftp_up_independently_final.rds")
} else {
  bam_ftp_up_independently_final = bam(
    formula  = logit ~ model + s(log_ftp, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_ftp_up_independently_final, "bam_ftp_up_independently_final.rds")
}
summary(bam_ftp_up_independently_final)

set.seed(964)
if (file.exists("bam_ftp_up_subword_final.rds")) {
  bam_ftp_up_subword_final = readRDS("bam_ftp_up_subword_final.rds")
} else {
  bam_ftp_up_subword_final = bam(
    formula  = logit ~ model + s(log_ftp, by = model) + s(verb_up, bs = 're'),
    data     = up_sub_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_ftp_up_subword_final, "bam_ftp_up_subword_final.rds")
}
summary(bam_ftp_up_subword_final)
```

### Effect of FTP across layers — by model (mgcv, normalized layers)

`layer_norm` is already computed above. We filter to non-NA FTP rows.

```{r}
up_indep_ftp   = up_independently %>% filter(!is.na(log_ftp))
up_subword_ftp = up_subword       %>% filter(!is.na(log_ftp))

set.seed(964)
if (file.exists("model_ftp_layer_up_independently.rds")) {
  model_ftp_layer_up_independently = readRDS("model_ftp_layer_up_independently.rds")
} else {
  model_ftp_layer_up_independently = bam(
    formula  = logit ~ model + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_ftp_layer_up_independently, "model_ftp_layer_up_independently.rds")
}
summary(model_ftp_layer_up_independently)

set.seed(964)
if (file.exists("model_ftp_layer_up_subword.rds")) {
  model_ftp_layer_up_subword = readRDS("model_ftp_layer_up_subword.rds")
} else {
  model_ftp_layer_up_subword = bam(
    formula  = logit ~ model + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_subword_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_ftp_layer_up_subword, "model_ftp_layer_up_subword.rds")
}
summary(model_ftp_layer_up_subword)
```

### FTP across layers — plots (by model)
```{r}
# Adapts plot_freq_by_layer_by_model for FTP
plot_ftp_by_layer_by_model = function(model_fit, data, condition_label) {
  layer_norm_seq = seq(0, 1, length.out = 50)
  log_ftp_seq    = seq(min(data$log_ftp, na.rm = TRUE), max(data$log_ftp, na.rm = TRUE), length.out = 50)
  ftp_lo         = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi         = quantile(data$log_ftp, 0.9, na.rm = TRUE)

  newdata_sp = expand.grid(layer_norm = layer_norm_seq, log_ftp = log_ftp_seq, model = levels(data$model)) %>%
    mutate(verb_up = data$verb_up[1])

  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit

  p_sp = newdata_sp %>%
    ggplot(aes(x = log_ftp, y = fit, color = layer_norm, group = interaction(model, layer_norm))) +
    geom_line(linewidth = 0.5, alpha = 0.7) +
    facet_wrap(~model) +
    scale_color_viridis_c(option = "plasma", name = "Layer\n(normalized)") +
    labs(title = paste0(condition_label, ": FTP effect by layer — by model"), x = "log(FTP)", y = "Predicted logit") +
    theme_minimal()

  newdata_diff = expand.grid(layer_norm = layer_norm_seq, model = levels(data$model), ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi), verb_up = data$verb_up[1])

  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE, exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit
  newdata_diff$se  = preds_diff$se.fit

  p_diff = newdata_diff %>%
    select(layer_norm, model, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff
    ) %>%
    ggplot(aes(x = layer_norm, y = diff, color = model, fill = model)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(condition_label, ": FTP effect (high - low) across layers — by model"),
      x = "Normalized layer (0 = first, 1 = final)", y = "Difference in predicted logit",
      color = "Model", fill = "Model"
    ) +
    theme_minimal()

  list(spaghetti = p_sp, diff = p_diff)
}

plots_ftp_indep   = plot_ftp_by_layer_by_model(model_ftp_layer_up_independently, up_indep_ftp,   "UP independently")
plots_ftp_subword = plot_ftp_by_layer_by_model(model_ftp_layer_up_subword,       up_subword_ftp, "UP as subword")

plots_ftp_indep$spaghetti
plots_ftp_indep$diff

plots_ftp_subword$spaghetti
plots_ftp_subword$diff
```

---

## Joint frequency + predictability

### Additive freq + FTP at the final layer — by model (brms)
```{r}
model_freq_ftp_up_independently = brm(
  formula = logit ~ log_freq * model + log_ftp * model + (1 | verb_up),
  data    = up_indep_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_ftp_up_independently",
  seed    = 964
)

model_freq_ftp_up_subword = brm(
  formula = logit ~ log_freq * model + log_ftp * model + (1 | verb_up),
  data    = up_sub_final_ftp,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_ftp_up_subword",
  seed    = 964
)
```

### Joint freq × FTP surface at final layer — by model (bam)
```{r}
set.seed(964)
if (file.exists("bam_joint_up_independently_final.rds")) {
  bam_joint_up_independently_final = readRDS("bam_joint_up_independently_final.rds")
} else {
  bam_joint_up_independently_final = bam(
    formula  = logit ~ model + te(log_freq, log_ftp, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_joint_up_independently_final, "bam_joint_up_independently_final.rds")
}
summary(bam_joint_up_independently_final)

set.seed(964)
if (file.exists("bam_joint_up_subword_final.rds")) {
  bam_joint_up_subword_final = readRDS("bam_joint_up_subword_final.rds")
} else {
  bam_joint_up_subword_final = bam(
    formula  = logit ~ model + te(log_freq, log_ftp, by = model) + s(verb_up, bs = 're'),
    data     = up_sub_final_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_joint_up_subword_final, "bam_joint_up_subword_final.rds")
}
summary(bam_joint_up_subword_final)
```

### Freq and FTP effects across layers — by model (additive bam)
```{r}
set.seed(964)
if (file.exists("model_freq_ftp_layer_up_independently.rds")) {
  model_freq_ftp_layer_up_independently = readRDS("model_freq_ftp_layer_up_independently.rds")
} else {
  model_freq_ftp_layer_up_independently = bam(
    formula  = logit ~ model + te(log_freq, layer_norm, by = model) + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_indep_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_ftp_layer_up_independently, "model_freq_ftp_layer_up_independently.rds")
}
summary(model_freq_ftp_layer_up_independently)

set.seed(964)
if (file.exists("model_freq_ftp_layer_up_subword.rds")) {
  model_freq_ftp_layer_up_subword = readRDS("model_freq_ftp_layer_up_subword.rds")
} else {
  model_freq_ftp_layer_up_subword = bam(
    formula  = logit ~ model + te(log_freq, layer_norm, by = model) + te(log_ftp, layer_norm, by = model) + s(verb_up, bs = 're'),
    data     = up_subword_ftp,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_ftp_layer_up_subword, "model_freq_ftp_layer_up_subword.rds")
}
summary(model_freq_ftp_layer_up_subword)
```
