---
title: "writeup"
format: html
editor: visual
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)

# ── Constants & helpers ────────────────────────────────────────────────────────
OLMO_CACHE_DIR      = "../model_cache/olmo"
BLM_CACHE_DIR       = "../model_cache/babylm"
WH_CACHE_DIR        = "../model_cache/whisper"

OLMO_FINAL_LAYER    = 31L
OLMO_N_LAYERS       = 32L
BLM_FINAL_LAYER_MAP = list("opt-125m" = 11L, "opt-350m" = 23L, "opt-1.3b" = 23L)
BLM_N_LAYERS        = c("opt-125m" = 12L, "opt-350m" = 24L, "opt-1.3b" = 24L)
WH_FINAL_LAYER      = 11L
WH_N_LAYERS         = 12L

BLM_TAGS      = c("opt-125m", "opt-350m", "opt-1.3b")
WH_COMPONENTS = c("encoder", "decoder")
blm_slug      = function(tag) gsub("\\.", "", gsub("opt-", "", tag))

BLM_LABELS = c("opt-125m" = "BabyLM 125M", "opt-350m" = "BabyLM 350M", "opt-1.3b" = "BabyLM 1.3B")
WH_LABELS  = c("encoder"  = "Encoder",      "decoder"  = "Decoder")

# Load a cached bam/RDS model
olmo_rds = function(name) readRDS(file.path(OLMO_CACHE_DIR, paste0(name, ".rds")))
blm_rds  = function(name) readRDS(file.path(BLM_CACHE_DIR,  paste0(name, ".rds")))
wh_rds   = function(name) readRDS(file.path(WH_CACHE_DIR,   paste0(name, ".rds")))

# brms file= paths (brms appends .rds itself)
olmo_brms_path = function(name) file.path(OLMO_CACHE_DIR, name)
blm_brms_path  = function(name) file.path(BLM_CACHE_DIR,  name)
wh_brms_path   = function(name) file.path(WH_CACHE_DIR,   name)

priors = c(prior(normal(0, 1), class = "b"))

# ── Raw data ───────────────────────────────────────────────────────────────────
olmo_indep = read_csv("../Data/olmo-3-7b/Data_up/all_layers_results.csv",
                       show_col_types = FALSE) %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp / (1 - ftp)),
         verb_up = factor(verb_up)) %>%
  filter(!is.na(ftp), is.finite(log_ftp))

olmo_sub = read_csv("../Data/olmo-3-7b/Data_upsubword/all_layers_results.csv",
                     show_col_types = FALSE) %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp / (1 - ftp)),
         verb_up = factor(verb_up)) %>%
  filter(!is.na(ftp), is.finite(log_ftp))

load_babylm_csvs = function(tag) {
  list(
    ui = read_csv(paste0("../Data/babylm/", tag, "/Data_up/all_layers_results.csv"),
                  show_col_types = FALSE) %>%
      mutate(model = tag, log_freq = log(frequency),
             log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up)) %>%
      filter(!is.na(ftp), is.finite(log_ftp)),
    us = read_csv(paste0("../Data/babylm/", tag, "/Data_upsubword/all_layers_results.csv"),
                  show_col_types = FALSE) %>%
      mutate(model = tag, log_freq = log(frequency),
             log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up)) %>%
      filter(!is.na(ftp), is.finite(log_ftp))
  )
}
blm_raw   = setNames(lapply(BLM_TAGS, load_babylm_csvs), BLM_TAGS)
blm_indep = map_dfr(blm_raw, "ui") %>%
  mutate(model = factor(model, levels = BLM_TAGS))
blm_sub   = map_dfr(blm_raw, "us") %>%
  mutate(model = factor(model, levels = BLM_TAGS))

ftp_lookup = read_csv("../Data/ftp_lookup.csv", show_col_types = FALSE)
load_whisper_component = function(path, comp) {
  read_csv(path, show_col_types = FALSE) %>%
    select(-any_of("ftp")) %>%
    mutate(component = comp, log_freq = log(frequency),
           verb_up_chr = as.character(verb_up)) %>%
    left_join(ftp_lookup, by = c("verb_up_chr" = "verb_up")) %>%
    filter(!is.na(ftp)) %>%
    mutate(log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up_chr)) %>%
    filter(is.finite(log_ftp)) %>% select(-verb_up_chr)
}
whisper_all = bind_rows(
  load_whisper_component("../Data/whisper_audio/encoder/all_layers_results.csv", "encoder"),
  load_whisper_component("../Data/whisper_audio/decoder/all_layers_results.csv", "decoder")
) %>% mutate(component = factor(component, levels = c("encoder", "decoder")))

# ── Item statistics ────────────────────────────────────────────────────────────
olmo_items = olmo_indep %>% filter(layer == 0) %>% group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

blm_items = blm_indep %>% filter(layer == 0, model == "opt-125m") %>% group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

whisper_items = whisper_all %>% filter(layer == 0, component == "encoder") %>%
  group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

# ── brms: frequency — first & final layer, UP independently & subword ─────────
# OLMo
olmo_brms_freq_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_independently.rds"))
olmo_brms_freq_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_independently_first_layer.rds"))
olmo_brms_freq_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_subword.rds"))
olmo_brms_freq_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_subword_first_layer.rds"))

# BabyLM (named lists by tag)
blm_brms_freq_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

# Whisper (named lists by component)
wh_brms_freq_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_freq_final_", comp, ".rds")))), WH_COMPONENTS)
wh_brms_freq_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_freq_first_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: FTP — first & final layer, UP independently & subword ───────────────
# OLMo
olmo_brms_ftp_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_independently.rds"))
olmo_brms_ftp_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_independently_first_layer.rds"))
olmo_brms_ftp_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_subword.rds"))
olmo_brms_ftp_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_subword_first_layer.rds"))

# BabyLM
blm_brms_ftp_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

# Whisper
wh_brms_ftp_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_ftp_final_", comp, ".rds")))), WH_COMPONENTS)
wh_brms_ftp_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_ftp_first_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: joint freq + FTP — final layer ──────────────────────────────────────
olmo_brms_joint_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_independently.rds"))
olmo_brms_joint_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_subword.rds"))

blm_brms_joint_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_joint_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)

wh_brms_joint_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_joint_final_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: joint freq + FTP — first layer ──────────────────────────────────────
olmo_brms_joint_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_independently_first.rds"))
olmo_brms_joint_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_subword_first.rds"))

blm_brms_joint_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_joint_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

wh_brms_joint_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_joint_first_", comp, ".rds")))), WH_COMPONENTS)

# ── GAM: frequency × layer ────────────────────────────────────────────────────
olmo_gam_freq_indep = olmo_rds("model_freq_layer_up_independently")
olmo_gam_freq_sub   = olmo_rds("model_freq_layer_up_subword")

blm_gam_freq_indep = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_freq_layer_indep_", blm_slug(tag)))), BLM_TAGS)
blm_gam_freq_sub   = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_freq_layer_sub_", blm_slug(tag)))), BLM_TAGS)

wh_gam_freq = setNames(lapply(WH_COMPONENTS, function(comp)
  wh_rds(paste0("model_freq_layer_", comp))), WH_COMPONENTS)

# ── GAM: FTP × layer ──────────────────────────────────────────────────────────
olmo_gam_ftp_indep = olmo_rds("model_ftp_layer_up_independently")
olmo_gam_ftp_sub   = olmo_rds("model_ftp_layer_up_subword")

blm_gam_ftp_indep = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_ftp_layer_indep_", blm_slug(tag)))), BLM_TAGS)
blm_gam_ftp_sub   = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_ftp_layer_sub_", blm_slug(tag)))), BLM_TAGS)

wh_gam_ftp = setNames(lapply(WH_COMPONENTS, function(comp)
  wh_rds(paste0("model_ftp_layer_", comp))), WH_COMPONENTS)

# ── brms table helpers ────────────────────────────────────────────────────────
brms_df = function(model, model_name) {
  fe     = as.data.frame(fixef(model))
  params = rownames(fe)
  rownames(fe) = NULL
  post = as_draws_df(model)
  pgt0 = sapply(params, function(p) {
    col = paste0("b_", p)
    if (col %in% names(post)) round(mean(post[[col]] > 0) * 100, 1) else NA_real_
  })
  fe %>%
    mutate(Model = model_name, Parameter = params, `% > 0` = pgt0) %>%
    select(Model, Parameter, Estimate, Est.Error, Q2.5, Q97.5, `% > 0`)
}

brms_combined_table = function(model_list) {
  bind_rows(mapply(brms_df, model_list, names(model_list), SIMPLIFY = FALSE)) %>%
    knitr::kable(digits = 3, row.names = FALSE)
}

# ── Diff data helpers (return data frames for faceted plotting) ────────────────
freq_diff_data = function(model, data, n_layers, panel) {
  layer_seq   = 0:(n_layers - 1)
  log_freq_lo = quantile(data$log_freq, 0.1)
  log_freq_hi = quantile(data$log_freq, 0.9)
  newdata = expand.grid(layer = layer_seq, freq_level = c("low", "high")) %>%
    mutate(log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
           verb_up  = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = TRUE, exclude = "s(verb_up)")
  newdata %>%
    mutate(fit = preds$fit, se = preds$se.fit) %>%
    select(layer, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(diff    = fit_high - fit_low,
           se_diff = sqrt(se_high^2 + se_low^2),
           lo95    = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80    = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff,
           panel   = panel)
}

ftp_diff_data = function(model, data, n_layers, panel) {
  layer_seq = 0:(n_layers - 1)
  ftp_lo    = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi    = quantile(data$log_ftp, 0.9, na.rm = TRUE)
  newdata = expand.grid(layer = layer_seq, ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi),
           verb_up = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = TRUE, exclude = "s(verb_up)")
  newdata %>%
    mutate(fit = preds$fit, se = preds$se.fit) %>%
    select(layer, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(diff    = fit_high - fit_low,
           se_diff = sqrt(se_high^2 + se_low^2),
           lo95    = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80    = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff,
           panel   = panel)
}

plot_freq_diff = function(df, ncol = 2) {
  df %>%
    mutate(panel = factor(panel, levels = unique(panel))) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    facet_wrap(~ panel, ncol = ncol, scales = "free_x") +
    labs(x = "Layer", y = expression(Delta ~ "logit (90th" ~ "-" ~ "10th pct freq)")) +
    theme_minimal()
}

plot_ftp_diff = function(df, ncol = 2) {
  df %>%
    mutate(panel = factor(panel, levels = unique(panel))) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "darkorange", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "darkorange", alpha = 0.3) +
    geom_line(linewidth = 1, color = "darkorange4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    facet_wrap(~ panel, ncol = ncol, scales = "free_x") +
    labs(x = "Layer", y = expression(Delta ~ "logit (90th" ~ "-" ~ "10th pct FTP)")) +
    theme_minimal()
}

# ── Diff data: frequency ───────────────────────────────────────────────────────
olmo_ddata_freq_indep = freq_diff_data(olmo_gam_freq_indep, olmo_indep, OLMO_N_LAYERS, "OLMo-3 7B")
olmo_ddata_freq_sub   = freq_diff_data(olmo_gam_freq_sub,   olmo_sub,   OLMO_N_LAYERS, "OLMo-3 7B")

blm_ddata_freq_indep = bind_rows(lapply(BLM_TAGS, function(tag)
  freq_diff_data(blm_gam_freq_indep[[tag]], blm_indep %>% filter(model == tag),
                 BLM_N_LAYERS[[tag]], BLM_LABELS[[tag]])))
blm_ddata_freq_sub   = bind_rows(lapply(BLM_TAGS, function(tag)
  freq_diff_data(blm_gam_freq_sub[[tag]], blm_sub %>% filter(model == tag),
                 BLM_N_LAYERS[[tag]], BLM_LABELS[[tag]])))

wh_ddata_freq = bind_rows(lapply(WH_COMPONENTS, function(comp)
  freq_diff_data(wh_gam_freq[[comp]], whisper_all %>% filter(component == comp),
                 WH_N_LAYERS, WH_LABELS[[comp]])))

# ── Diff data: FTP ─────────────────────────────────────────────────────────────
olmo_ddata_ftp_indep = ftp_diff_data(olmo_gam_ftp_indep, olmo_indep, OLMO_N_LAYERS, "OLMo-3 7B")
olmo_ddata_ftp_sub   = ftp_diff_data(olmo_gam_ftp_sub,   olmo_sub,   OLMO_N_LAYERS, "OLMo-3 7B")

blm_ddata_ftp_indep = bind_rows(lapply(BLM_TAGS, function(tag)
  ftp_diff_data(blm_gam_ftp_indep[[tag]], blm_indep %>% filter(model == tag),
                BLM_N_LAYERS[[tag]], BLM_LABELS[[tag]])))
blm_ddata_ftp_sub   = bind_rows(lapply(BLM_TAGS, function(tag)
  ftp_diff_data(blm_gam_ftp_sub[[tag]], blm_sub %>% filter(model == tag),
                BLM_N_LAYERS[[tag]], BLM_LABELS[[tag]])))

wh_ddata_ftp = bind_rows(lapply(WH_COMPONENTS, function(comp)
  ftp_diff_data(wh_gam_ftp[[comp]], whisper_all %>% filter(component == comp),
                WH_N_LAYERS, WH_LABELS[[comp]])))
```



# Experiment 1: UP Independently

## Items

The classifier was trained to distinguish the hidden-state representation of the particle *up* from representations of other tokens in the same sentence. Positive training examples consisted of 1,000 occurrences of standalone *up* (i.e., functioning as a verbal particle, as in *pick up*), drawn from sentences in the C4 corpus. Negative examples were 1,000 tokens randomly selected from the same sentences, restricted to tokens whose decoded string consists entirely of alphabetic characters (no numbers, punctuation, or special characters), and excluding *up* itself and any token containing *up* as a substring. The validation set was constructed identically (1,000 positive, 1,000 negative) from a non-overlapping set of sentences. Together, training and validation used approximately 2,000 sentences each.

The classifiers targeting OLMo-3 7B and the BabyLM models were trained and tested on the **same underlying sentences**, but the frequency and predictability statistics used in the analyses were derived from the respective corpus for each model (the C4 corpus for the OLMo model and the babyLM corpus for the babyLM models). In other words, the training and validation data for the models was identical, but the statistics used in the analysis varied.

The test set comprised verb–*up* phrasal verb types (e.g., *pick up*, *set up*) that occurred at least 10 times in the corpus. For each type, up to 20 sentences were sampled, yielding approximately 5,000 unique verb–*up* types and ~100,000 test instances in total. Corpus frequency and forward transitional probability (FTP; $P(\text{up} \mid V)$) were derived from the same corpus used to select training sentences. Item-level statistics for the test set are shown in @tbl-exp1-items.

```{r}
#| label: tbl-exp1-items
#| tbl-cap: "Summary statistics for verb–up types in the OLMo-3 7B test set."
olmo_items %>%
  summarise(
    N = n(),
    `Mean freq` = round(mean(frequency), 1),
    `SD freq`   = round(sd(frequency),   1),
    `Mean FTP`  = round(mean(ftp, na.rm = TRUE), 3),
    `SD FTP`    = round(sd(ftp,   na.rm = TRUE), 3)
  ) %>%
  knitr::kable(row.names = FALSE)
```

## Joint model — first layer

```{r}
#| label: tbl-exp1-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_indep_first,
  "BabyLM 125M" = blm_brms_joint_indep_first[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_indep_first[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_indep_first[["opt-1.3b"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp1-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_indep_final,
  "BabyLM 125M" = blm_brms_joint_indep_final[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_indep_final[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_indep_final[["opt-1.3b"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp1-freq
#| fig-cap: "Estimated frequency effect (difference in predicted logit between the 90th and 10th percentile of log frequency) across layers for UP independently. Ribbons show 80% and 95% confidence intervals. Panels show OLMo-3 7B and BabyLM models of increasing size."
bind_rows(olmo_ddata_freq_indep, blm_ddata_freq_indep) %>% plot_freq_diff()
```

## FTP effect across layers

```{r}
#| label: fig-exp1-ftp
#| fig-cap: "Estimated FTP effect (difference in predicted logit between the 90th and 10th percentile of log-odds FTP) across layers for UP independently. Ribbons show 80% and 95% confidence intervals. Panels show OLMo-3 7B and BabyLM models of increasing size."
bind_rows(olmo_ddata_ftp_indep, blm_ddata_ftp_indep) %>% plot_ftp_diff()
```

# Experiment 2: UP as Subword

## Items

The training set for this experiment combined two types of positive example: 1,000 occurrences of standalone *up* (identical to Experiment 1) and 1,000 occurrences of *up* embedded within a larger word (e.g., *update*, *upon*, *setup*). Critically, the subword positives were restricted to **unique word types**: each up-containing word form contributed exactly one instance, so the classifier could not learn to recognise a high-frequency form like *setup* from repeated exposure. Negative examples (1,000 drawn from each sentence pool) were tokens consisting entirely of alphabetic characters that did not contain *up* as a substring. The validation set was constructed identically. As in Experiment 1, OLMo-3 7B and BabyLM models used the same underlying sentences with their own tokenizers, and BabyLM models used BabyLM-corpus statistics for frequency and FTP. The test set was identical to Experiment 1: all verb–*up* phrasal verb types with at least 10 corpus occurrences, sampled at up to 20 sentences per type. Item-level statistics are shown in @tbl-exp2-items.

```{r}
#| label: tbl-exp2-items
#| tbl-cap: "Summary statistics for verb–up types in the OLMo-3 7B test set (UP as subword condition; same items as Experiment 1)."
olmo_items %>%
  summarise(
    N = n(),
    `Mean freq` = round(mean(frequency), 1),
    `SD freq`   = round(sd(frequency),   1),
    `Mean FTP`  = round(mean(ftp, na.rm = TRUE), 3),
    `SD FTP`    = round(sd(ftp,   na.rm = TRUE), 3)
  ) %>%
  knitr::kable(row.names = FALSE)
```

## Joint model — first layer

```{r}
#| label: tbl-exp2-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_sub_first,
  "BabyLM 125M" = blm_brms_joint_sub_first[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_sub_first[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_sub_first[["opt-1.3b"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp2-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_sub_final,
  "BabyLM 125M" = blm_brms_joint_sub_final[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_sub_final[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_sub_final[["opt-1.3b"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp2-freq
#| fig-cap: "Estimated frequency effect (difference in predicted logit between the 90th and 10th percentile of log frequency) across layers for UP as subword. Ribbons show 80% and 95% confidence intervals. Panels show OLMo-3 7B and BabyLM models of increasing size."
bind_rows(olmo_ddata_freq_sub, blm_ddata_freq_sub) %>% plot_freq_diff()
```

## FTP effect across layers

```{r}
#| label: fig-exp2-ftp
#| fig-cap: "Estimated FTP effect (difference in predicted logit between the 90th and 10th percentile of log-odds FTP) across layers for UP as subword. Ribbons show 80% and 95% confidence intervals. Panels show OLMo-3 7B and BabyLM models of increasing size."
bind_rows(olmo_ddata_ftp_sub, blm_ddata_ftp_sub) %>% plot_ftp_diff()
```

# Experiment 3: Whisper (ASR Model)

## Items

The Whisper experiment used spoken language data drawn from a mixed-source audio metadata file containing timestamped speech segments in which a word immediately precedes *up*. Word-level timestamps within each segment were obtained using WhisperX forced alignment. Each occurrence of *up* was classified by running spaCy part-of-speech tagging on the full segment transcript: if the token immediately preceding *up* was tagged as a VERB, the instance was labelled *V+up* (e.g., *pick up*, *clean up*); otherwise it was labelled *word_up* (e.g., *it up*, *them up*). This classification procedure mirrors the one used to identify standalone *up* in Experiments 1 and 2.

The **training and validation sets** used 1,000 *word_up* instances each as positive examples, paired with 1,000 randomly selected non-*up* words from the same segment as negative examples. This mirrors the standalone-*up* training design of Experiments 1 and 2. Negative words were restricted to those whose string consists entirely of alphabetic characters (matching the `token.isalpha()` criterion used in Experiments 1 and 2), with a minimum duration of 20 ms. Training and validation segments were non-overlapping.

The **test set** comprised V+up phrasal verb types with at least 5 occurrences in the audio dataset, with up to 20 instances per qualifying type. The minimum-occurrence threshold is lower than in Experiments 1 and 2 (which required at least 10 corpus sentences per type), reflecting the sparser coverage of audio relative to written text. Corpus frequency and FTP values were drawn from the C4-derived lookup table used in Experiments 1 and 2. Item-level statistics for the Whisper test set are shown in @tbl-exp3-items.

```{r}
#| label: tbl-exp3-items
#| tbl-cap: "Summary statistics for verb–up types in the Whisper test set."
whisper_items %>%
  summarise(
    N = n(),
    `Mean freq` = round(mean(frequency), 1),
    `SD freq`   = round(sd(frequency),   1),
    `Mean FTP`  = round(mean(ftp, na.rm = TRUE), 3),
    `SD FTP`    = round(sd(ftp,   na.rm = TRUE), 3)
  ) %>%
  knitr::kable(row.names = FALSE)
```

## Joint model — first layer

```{r}
#| label: tbl-exp3-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "Encoder" = wh_brms_joint_first[["encoder"]],
  "Decoder" = wh_brms_joint_first[["decoder"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp3-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "Encoder" = wh_brms_joint_final[["encoder"]],
  "Decoder" = wh_brms_joint_final[["decoder"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp3-freq
#| fig-cap: "Estimated frequency effect across layers for Whisper encoder and decoder. Ribbons show 80% and 95% confidence intervals."
wh_ddata_freq %>% plot_freq_diff(ncol = 2)
```

## FTP effect across layers

```{r}
#| label: fig-exp3-ftp
#| fig-cap: "Estimated FTP effect across layers for Whisper encoder and decoder. Ribbons show 80% and 95% confidence intervals."
wh_ddata_ftp %>% plot_ftp_diff(ncol = 2)
```
