---
title: "writeup"
format: html
editor: visual
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)

# ── Constants & helpers ────────────────────────────────────────────────────────
OLMO_CACHE_DIR      = "../model_cache/olmo"
BLM_CACHE_DIR       = "../model_cache/babylm"
WH_CACHE_DIR        = "../model_cache/whisper"

OLMO_FINAL_LAYER    = 31L
OLMO_N_LAYERS       = 32L
BLM_FINAL_LAYER_MAP = list("opt-125m" = 11L, "opt-350m" = 23L, "opt-1.3b" = 23L)
BLM_N_LAYERS        = c("opt-125m" = 12L, "opt-350m" = 24L, "opt-1.3b" = 24L)
WH_FINAL_LAYER      = 11L
WH_N_LAYERS         = 12L

BLM_TAGS      = c("opt-125m", "opt-350m", "opt-1.3b")
WH_COMPONENTS = c("encoder", "decoder")
blm_slug      = function(tag) gsub("\\.", "", gsub("opt-", "", tag))

# Load a cached bam/RDS model
olmo_rds = function(name) readRDS(file.path(OLMO_CACHE_DIR, paste0(name, ".rds")))
blm_rds  = function(name) readRDS(file.path(BLM_CACHE_DIR,  paste0(name, ".rds")))
wh_rds   = function(name) readRDS(file.path(WH_CACHE_DIR,   paste0(name, ".rds")))

# brms file= paths (brms appends .rds itself)
olmo_brms_path = function(name) file.path(OLMO_CACHE_DIR, name)
blm_brms_path  = function(name) file.path(BLM_CACHE_DIR,  name)
wh_brms_path   = function(name) file.path(WH_CACHE_DIR,   name)

priors = c(prior(normal(0, 1), class = "b"))

# ── Raw data ───────────────────────────────────────────────────────────────────
olmo_indep = read_csv("../Data/olmo-3-7b/Data_up/all_layers_results.csv",
                       show_col_types = FALSE) %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp / (1 - ftp)),
         verb_up = factor(verb_up)) %>%
  filter(!is.na(ftp), is.finite(log_ftp))

olmo_sub = read_csv("../Data/olmo-3-7b/Data_upsubword/all_layers_results.csv",
                     show_col_types = FALSE) %>%
  mutate(log_freq = log(frequency), log_ftp = log(ftp / (1 - ftp)),
         verb_up = factor(verb_up)) %>%
  filter(!is.na(ftp), is.finite(log_ftp))

load_babylm_csvs = function(tag) {
  list(
    ui = read_csv(paste0("../Data/babylm/", tag, "/Data_up/all_layers_results.csv"),
                  show_col_types = FALSE) %>%
      mutate(model = tag, log_freq = log(frequency),
             log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up)) %>%
      filter(!is.na(ftp), is.finite(log_ftp)),
    us = read_csv(paste0("../Data/babylm/", tag, "/Data_upsubword/all_layers_results.csv"),
                  show_col_types = FALSE) %>%
      mutate(model = tag, log_freq = log(frequency),
             log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up)) %>%
      filter(!is.na(ftp), is.finite(log_ftp))
  )
}
blm_raw   = setNames(lapply(BLM_TAGS, load_babylm_csvs), BLM_TAGS)
blm_indep = map_dfr(blm_raw, "ui") %>%
  mutate(model = factor(model, levels = BLM_TAGS))
blm_sub   = map_dfr(blm_raw, "us") %>%
  mutate(model = factor(model, levels = BLM_TAGS))

ftp_lookup = read_csv("../Data/ftp_lookup.csv", show_col_types = FALSE)
load_whisper_component = function(path, comp) {
  read_csv(path, show_col_types = FALSE) %>%
    select(-any_of("ftp")) %>%
    mutate(component = comp, log_freq = log(frequency),
           verb_up_chr = as.character(verb_up)) %>%
    left_join(ftp_lookup, by = c("verb_up_chr" = "verb_up")) %>%
    filter(!is.na(ftp)) %>%
    mutate(log_ftp = log(ftp / (1 - ftp)), verb_up = factor(verb_up_chr)) %>%
    filter(is.finite(log_ftp)) %>% select(-verb_up_chr)
}
whisper_all = bind_rows(
  load_whisper_component("../Data/whisper/encoder/all_layers_results.csv", "encoder"),
  load_whisper_component("../Data/whisper/decoder/all_layers_results.csv", "decoder")
) %>% mutate(component = factor(component, levels = c("encoder", "decoder")))

# ── Item statistics ────────────────────────────────────────────────────────────
olmo_items = olmo_indep %>% filter(layer == 0) %>% group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

blm_items = blm_indep %>% filter(layer == 0, model == "opt-125m") %>% group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

whisper_items = whisper_all %>% filter(layer == 0, component == "encoder") %>%
  group_by(verb_up) %>%
  summarise(frequency = first(frequency), ftp = first(ftp),
            log_freq = first(log_freq), log_ftp = first(log_ftp), .groups = "drop")

# ── brms: frequency — first & final layer, UP independently & subword ─────────
# OLMo
olmo_brms_freq_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_independently.rds"))
olmo_brms_freq_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_independently_first_layer.rds"))
olmo_brms_freq_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_subword.rds"))
olmo_brms_freq_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_up_subword_first_layer.rds"))

# BabyLM (named lists by tag)
blm_brms_freq_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_freq_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_freq_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

# Whisper (named lists by component)
wh_brms_freq_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_freq_final_", comp, ".rds")))), WH_COMPONENTS)
wh_brms_freq_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_freq_first_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: FTP — first & final layer, UP independently & subword ───────────────
# OLMo
olmo_brms_ftp_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_independently.rds"))
olmo_brms_ftp_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_independently_first_layer.rds"))
olmo_brms_ftp_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_subword.rds"))
olmo_brms_ftp_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_ftp_up_subword_first_layer.rds"))

# BabyLM
blm_brms_ftp_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_ftp_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_ftp_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

# Whisper
wh_brms_ftp_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_ftp_final_", comp, ".rds")))), WH_COMPONENTS)
wh_brms_ftp_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_ftp_first_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: joint freq + FTP — final layer ──────────────────────────────────────
olmo_brms_joint_indep_final = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_independently.rds"))
olmo_brms_joint_sub_final   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_subword.rds"))

blm_brms_joint_indep_final = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_indep_final_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_joint_sub_final   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_sub_final_", blm_slug(tag), ".rds")))), BLM_TAGS)

wh_brms_joint_final = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_joint_final_", comp, ".rds")))), WH_COMPONENTS)

# ── brms: joint freq + FTP — first layer ──────────────────────────────────────
olmo_brms_joint_indep_first = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_independently_first.rds"))
olmo_brms_joint_sub_first   = readRDS(file.path(OLMO_CACHE_DIR, "model_freq_ftp_up_subword_first.rds"))

blm_brms_joint_indep_first = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_indep_first_", blm_slug(tag), ".rds")))), BLM_TAGS)
blm_brms_joint_sub_first   = setNames(lapply(BLM_TAGS, function(tag)
  readRDS(file.path(BLM_CACHE_DIR, paste0("model_joint_sub_first_", blm_slug(tag), ".rds")))), BLM_TAGS)

wh_brms_joint_first = setNames(lapply(WH_COMPONENTS, function(comp)
  readRDS(file.path(WH_CACHE_DIR, paste0("model_joint_first_", comp, ".rds")))), WH_COMPONENTS)

# ── GAM: frequency × layer ────────────────────────────────────────────────────
olmo_gam_freq_indep = olmo_rds("model_freq_layer_up_independently")
olmo_gam_freq_sub   = olmo_rds("model_freq_layer_up_subword")

blm_gam_freq_indep = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_freq_layer_indep_", blm_slug(tag)))), BLM_TAGS)
blm_gam_freq_sub   = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_freq_layer_sub_", blm_slug(tag)))), BLM_TAGS)

wh_gam_freq = setNames(lapply(WH_COMPONENTS, function(comp)
  wh_rds(paste0("model_freq_layer_", comp))), WH_COMPONENTS)

# ── GAM: FTP × layer ──────────────────────────────────────────────────────────
olmo_gam_ftp_indep = olmo_rds("model_ftp_layer_up_independently")
olmo_gam_ftp_sub   = olmo_rds("model_ftp_layer_up_subword")

blm_gam_ftp_indep = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_ftp_layer_indep_", blm_slug(tag)))), BLM_TAGS)
blm_gam_ftp_sub   = setNames(lapply(BLM_TAGS, function(tag)
  blm_rds(paste0("model_ftp_layer_sub_", blm_slug(tag)))), BLM_TAGS)

wh_gam_ftp = setNames(lapply(WH_COMPONENTS, function(comp)
  wh_rds(paste0("model_ftp_layer_", comp))), WH_COMPONENTS)

# ── brms table helpers ────────────────────────────────────────────────────────
brms_df = function(model, model_name) {
  fe   = as.data.frame(fixef(model))
  post = as_draws_df(model)
  pgt0 = sapply(rownames(fe), function(p) {
    col = paste0("b_", p)
    if (col %in% names(post)) round(mean(post[[col]] > 0) * 100, 1) else NA_real_
  })
  fe %>%
    mutate(Model = model_name, Parameter = rownames(fe), `% > 0` = pgt0) %>%
    select(Model, Parameter, Estimate, Est.Error, Q2.5, Q97.5, `% > 0`)
}

brms_combined_table = function(model_list) {
  bind_rows(mapply(brms_df, model_list, names(model_list), SIMPLIFY = FALSE)) %>%
    knitr::kable(digits = 3)
}

# ── Diff plot helper functions ─────────────────────────────────────────────────
make_freq_diff_plot = function(model, data, n_layers) {
  layer_seq   = 0:(n_layers - 1)
  log_freq_lo = quantile(data$log_freq, 0.1)
  log_freq_hi = quantile(data$log_freq, 0.9)
  newdata = expand.grid(layer = layer_seq, freq_level = c("low", "high")) %>%
    mutate(log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
           verb_up  = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = TRUE, exclude = "s(verb_up)")
  newdata %>%
    mutate(fit = preds$fit, se = preds$se.fit) %>%
    select(layer, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(diff  = fit_high - fit_low,
           se_diff = sqrt(se_high^2 + se_low^2),
           lo95  = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80  = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
    geom_line(linewidth = 1, color = "steelblue4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(x = "Layer", y = "Δ logit (90th − 10th pct freq)") +
    theme_minimal()
}

make_ftp_diff_plot = function(model, data, n_layers) {
  layer_seq = 0:(n_layers - 1)
  ftp_lo    = quantile(data$log_ftp, 0.1, na.rm = TRUE)
  ftp_hi    = quantile(data$log_ftp, 0.9, na.rm = TRUE)
  newdata = expand.grid(layer = layer_seq, ftp_level = c("low", "high")) %>%
    mutate(log_ftp = ifelse(ftp_level == "low", ftp_lo, ftp_hi),
           verb_up = data$verb_up[1])
  preds = predict(model, newdata = newdata, se.fit = TRUE, exclude = "s(verb_up)")
  newdata %>%
    mutate(fit = preds$fit, se = preds$se.fit) %>%
    select(layer, ftp_level, fit, se) %>%
    pivot_wider(names_from = ftp_level, values_from = c(fit, se)) %>%
    mutate(diff    = fit_high - fit_low,
           se_diff = sqrt(se_high^2 + se_low^2),
           lo95    = diff - 1.96 * se_diff, hi95 = diff + 1.96 * se_diff,
           lo80    = diff - 1.28 * se_diff, hi80 = diff + 1.28 * se_diff) %>%
    ggplot(aes(x = layer, y = diff)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "darkorange", alpha = 0.2) +
    geom_ribbon(aes(ymin = lo80, ymax = hi80), fill = "darkorange", alpha = 0.3) +
    geom_line(linewidth = 1, color = "darkorange4") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(x = "Layer", y = "Δ logit (90th − 10th pct FTP)") +
    theme_minimal()
}

# ── Diff plots: frequency ──────────────────────────────────────────────────────
# OLMo
olmo_diff_freq_indep = make_freq_diff_plot(olmo_gam_freq_indep, olmo_indep, OLMO_N_LAYERS)
olmo_diff_freq_sub   = make_freq_diff_plot(olmo_gam_freq_sub,   olmo_sub,   OLMO_N_LAYERS)

# BabyLM (named lists)
blm_diff_freq_indep = setNames(lapply(BLM_TAGS, function(tag)
  make_freq_diff_plot(blm_gam_freq_indep[[tag]], blm_indep %>% filter(model == tag),
                      BLM_N_LAYERS[[tag]])), BLM_TAGS)
blm_diff_freq_sub   = setNames(lapply(BLM_TAGS, function(tag)
  make_freq_diff_plot(blm_gam_freq_sub[[tag]], blm_sub %>% filter(model == tag),
                      BLM_N_LAYERS[[tag]])), BLM_TAGS)

# Whisper (named lists by component)
wh_diff_freq = setNames(lapply(WH_COMPONENTS, function(comp)
  make_freq_diff_plot(wh_gam_freq[[comp]], whisper_all %>% filter(component == comp),
                      WH_N_LAYERS)), WH_COMPONENTS)

# ── Diff plots: FTP ────────────────────────────────────────────────────────────
# OLMo
olmo_diff_ftp_indep = make_ftp_diff_plot(olmo_gam_ftp_indep, olmo_indep, OLMO_N_LAYERS)
olmo_diff_ftp_sub   = make_ftp_diff_plot(olmo_gam_ftp_sub,   olmo_sub,   OLMO_N_LAYERS)

# BabyLM
blm_diff_ftp_indep = setNames(lapply(BLM_TAGS, function(tag)
  make_ftp_diff_plot(blm_gam_ftp_indep[[tag]], blm_indep %>% filter(model == tag),
                     BLM_N_LAYERS[[tag]])), BLM_TAGS)
blm_diff_ftp_sub   = setNames(lapply(BLM_TAGS, function(tag)
  make_ftp_diff_plot(blm_gam_ftp_sub[[tag]], blm_sub %>% filter(model == tag),
                     BLM_N_LAYERS[[tag]])), BLM_TAGS)

# Whisper
wh_diff_ftp = setNames(lapply(WH_COMPONENTS, function(comp)
  make_ftp_diff_plot(wh_gam_ftp[[comp]], whisper_all %>% filter(component == comp),
                     WH_N_LAYERS)), WH_COMPONENTS)
```



# Experiment 1: UP Independently

## Joint model — first layer

```{r}
#| label: tbl-exp1-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_indep_first,
  "BabyLM 125M" = blm_brms_joint_indep_first[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_indep_first[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_indep_first[["opt-1.3b"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp1-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_indep_final,
  "BabyLM 125M" = blm_brms_joint_indep_final[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_indep_final[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_indep_final[["opt-1.3b"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp1-freq
#| fig-cap: "Estimated frequency effect (difference in predicted logit between the 90th and 10th percentile of log frequency) across layers for UP independently. Ribbons show 80% and 95% credible intervals."
(olmo_diff_freq_indep | blm_diff_freq_indep[["opt-125m"]]) /
(blm_diff_freq_indep[["opt-350m"]] | blm_diff_freq_indep[["opt-1.3b"]])
```

## FTP effect across layers

```{r}
#| label: fig-exp1-ftp
#| fig-cap: "Estimated FTP effect (difference in predicted logit between the 90th and 10th percentile of log-odds FTP) across layers for UP independently. Ribbons show 80% and 95% credible intervals."
(olmo_diff_ftp_indep | blm_diff_ftp_indep[["opt-125m"]]) /
(blm_diff_ftp_indep[["opt-350m"]] | blm_diff_ftp_indep[["opt-1.3b"]])
```

# Experiment 2: UP as Subword

## Joint model — first layer

```{r}
#| label: tbl-exp2-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_sub_first,
  "BabyLM 125M" = blm_brms_joint_sub_first[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_sub_first[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_sub_first[["opt-1.3b"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp2-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "OLMo-3 7B"   = olmo_brms_joint_sub_final,
  "BabyLM 125M" = blm_brms_joint_sub_final[["opt-125m"]],
  "BabyLM 350M" = blm_brms_joint_sub_final[["opt-350m"]],
  "BabyLM 1.3B" = blm_brms_joint_sub_final[["opt-1.3b"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp2-freq
#| fig-cap: "Estimated frequency effect (difference in predicted logit between the 90th and 10th percentile of log frequency) across layers for UP as subword. Ribbons show 80% and 95% credible intervals."
(olmo_diff_freq_sub | blm_diff_freq_sub[["opt-125m"]]) /
(blm_diff_freq_sub[["opt-350m"]] | blm_diff_freq_sub[["opt-1.3b"]])
```

## FTP effect across layers

```{r}
#| label: fig-exp2-ftp
#| fig-cap: "Estimated FTP effect (difference in predicted logit between the 90th and 10th percentile of log-odds FTP) across layers for UP as subword. Ribbons show 80% and 95% credible intervals."
(olmo_diff_ftp_sub | blm_diff_ftp_sub[["opt-125m"]]) /
(blm_diff_ftp_sub[["opt-350m"]] | blm_diff_ftp_sub[["opt-1.3b"]])
```

# Experiment 3: Whisper (ASR Model)

## Joint model — first layer

```{r}
#| label: tbl-exp3-joint-first
#| tbl-cap: "Joint frequency × FTP model at the first layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "Encoder" = wh_brms_joint_first[["encoder"]],
  "Decoder" = wh_brms_joint_first[["decoder"]]
))
```

## Joint model — final layer

```{r}
#| label: tbl-exp3-joint-final
#| tbl-cap: "Joint frequency × FTP model at the final layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
brms_combined_table(list(
  "Encoder" = wh_brms_joint_final[["encoder"]],
  "Decoder" = wh_brms_joint_final[["decoder"]]
))
```

## Frequency effect across layers

```{r}
#| label: fig-exp3-freq
#| fig-cap: "Estimated frequency effect across layers for Whisper encoder and decoder. Ribbons show 80% and 95% credible intervals."
wh_diff_freq[["encoder"]] | wh_diff_freq[["decoder"]]
```

## FTP effect across layers

```{r}
#| label: fig-exp3-ftp
#| fig-cap: "Estimated FTP effect across layers for Whisper encoder and decoder. Ribbons show 80% and 95% credible intervals."
wh_diff_ftp[["encoder"]] | wh_diff_ftp[["decoder"]]
```