---
title: "writeup"
acl-mode: review
format:
  acl-pdf:
    keep-tex: true
  html: default
editor: visual
author:
  - name:
      literal: Zachary Houghton
    email: znhoughton@ucdavis.edu
    affiliations:
      - name: University of California, Davis
  - name:
      literal: Kenji Sagae
    affiliations:
      - name: University of California, Davis
  - name:
      literal: Emily Morgan
    affiliations:
      - name: University of California, Davis
abstract: |
  (Abstract goes here.)
bibliography: references.bib
tbl-cap-location: bottom
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(patchwork)
library(kableExtra)

# ── Results directory ──────────────────────────────────────────────────────────
RES = "results"

# ── Load pre-computed CSVs ─────────────────────────────────────────────────────
olmo_items   = read_csv(file.path(RES, "olmo_items.csv"),   show_col_types = FALSE)
blm_items    = read_csv(file.path(RES, "blm_items.csv"),    show_col_types = FALSE)
wh_splits    = read_csv(file.path(RES, "wh_splits.csv"),    show_col_types = FALSE)
wh_items_by_comp = read_csv(file.path(RES, "whisper_items_by_component.csv"), show_col_types = FALSE)

tbl_exp1_joint_first = read_csv(file.path(RES, "tbl_exp1_joint_first.csv"), show_col_types = FALSE)
tbl_exp1_joint_final = read_csv(file.path(RES, "tbl_exp1_joint_final.csv"), show_col_types = FALSE)
tbl_exp2_joint_first = read_csv(file.path(RES, "tbl_exp2_joint_first.csv"), show_col_types = FALSE)
tbl_exp2_joint_final = read_csv(file.path(RES, "tbl_exp2_joint_final.csv"), show_col_types = FALSE)
tbl_exp3_joint_first = read_csv(file.path(RES, "tbl_exp3_joint_first.csv"), show_col_types = FALSE)
tbl_exp3_joint_final = read_csv(file.path(RES, "tbl_exp3_joint_final.csv"), show_col_types = FALSE)

surface_exp1_joint_olmo = read_csv(file.path(RES, "surface_exp1_joint_olmo.csv"), show_col_types = FALSE)
surface_exp1_joint_blm  = read_csv(file.path(RES, "surface_exp1_joint_blm.csv"),  show_col_types = FALSE)
surface_exp2_joint_olmo = read_csv(file.path(RES, "surface_exp2_joint_olmo.csv"), show_col_types = FALSE)
surface_exp2_joint_blm  = read_csv(file.path(RES, "surface_exp2_joint_blm.csv"),  show_col_types = FALSE)
surface_exp3_joint_wh   = read_csv(file.path(RES, "surface_exp3_joint_wh.csv"),   show_col_types = FALSE)
babylm_train_log        = read_csv(file.path(RES, "babylm_train_log.csv"),        show_col_types = FALSE)

# ── Plot helpers ───────────────────────────────────────────────────────────────
uo_colors = c("#007030", "#C9A92C", "#104735")

clean_brms_table = function(df) {
  param_labels = c(
    "Intercept"                = "Intercept",
    "c_log_freq"               = "Freq",
    "c_log_predic"             = "Predic.",
    "Ic_log_freqE2"            = "Freq\u00b2",
    "Ic_log_predicE2"          = "Predic.\u00b2",
    "c_log_freq:c_log_predic"  = "Freq:Predic."
  )
  df %>%
    mutate(
      Parameter = dplyr::recode(Parameter, !!!param_labels),
      `95% CI`  = sprintf("[%.2f, %.2f]", Q2.5, Q97.5),
      Estimate  = round(Estimate, 2),
      Est.Error = round(Est.Error, 2),
      `% > 0`   = round(`% > 0`, 1)
    ) %>%
    select(Model, Coef = Parameter, Est. = Estimate, `Err.` = Est.Error,
           `95% CI`, `% > 0`)
}

plot_marginal_surfaces = function(df) {
  df %>%
    mutate(label = factor(label, levels = unique(label))) %>%
    group_by(label) %>%
    group_modify(function(sub, key) {
      predic_mid = sub$c_log_predic[which.min(abs(sub$c_log_predic))]
      freq_mid   = sub$c_log_freq[which.min(abs(sub$c_log_freq))]
      bind_rows(
        sub %>% filter(abs(c_log_predic - predic_mid) < 1e-9) %>%
          transmute(x = c_log_freq   * freq_sd   + freq_mean,   fit, panel = "Frequency"),
        sub %>% filter(abs(c_log_freq - freq_mid)    < 1e-9) %>%
          transmute(x = c_log_predic * predic_sd + predic_mean, fit, panel = "Predictability")
      )
    }) %>%
    ungroup() %>%
    mutate(panel = factor(panel, levels = c("Frequency", "Predictability"))) %>%
    ggplot(aes(x = x, y = fit, color = label)) +
    geom_line(linewidth = 1) +
    facet_wrap(~ panel, scales = "free_x") +
    scale_color_manual(values = uo_colors) +
    labs(x = NULL, y = "Predicted logit", color = NULL) +
    theme_minimal() +
    theme(legend.position = "right")
}

blm_strip = function(df, layer) {
  df %>% filter(str_detect(label, layer)) %>%
    mutate(label = str_remove(label, " \u2014.*$"))
}

wh_strip = function(df, layer) {
  df %>% filter(str_detect(label, layer)) %>%
    mutate(label = str_remove(label, " \u2014.*$"))
}
```

# Model Training

We trained three autoregressive language models on the BabyLM corpus [@hu-etal-2024-findings], a 150M-token dataset designed to be more reflective of human language experience than web-scale corpora. All three models follow the OPT decoder-only transformer architecture, with 125M, 350M, and 1.3B parameters, respectively.

Prior to training, we fit a byte-pair encoding (BPE) tokenizer directly on the BabyLM training corpus, yielding a vocabulary of 8,192 subword types. This tokenizer was shared across all three model sizes, ensuring that cross-model comparisons are not confounded by differences in tokenization.

Models were trained from random initialization for 20 epochs using the AdamW optimizer with fused weight updates and bfloat16 mixed precision. The learning rate was set to $3 \times 10^{-4}$ for the 125M model and $1 \times 10^{-4}$ for the 350M and 1.3B models, each preceded by a linear warmup over 10% of total training steps. Training was distributed across two NVIDIA A100 80GB GPUs. Full hyperparameter details are given in @tbl-hyperparams (Appendix).

All models and training code are openly available on HuggingFace at `znhoughton/opt-babylm-{125m,350m,1.3b}-20eps-seed964`, including intermediate checkpoints saved every 20M tokens. Training loss curves and final validation perplexity are reported in @fig-train-loss, @tbl-val-perplexity, and @fig-val-perplexity (Appendix).

# Experiment 1: UP Independently

## Items

Placeholder reference: @houghton2020perceptualconsequencespauses.

The classifier was trained to distinguish the hidden-state representation of the preposition *up* from representations of other tokens in the same sentence. Positive training examples consisted of 1,000 occurrences of standalone *up* (i.e., functioning as a verbal particle, as in *pick up*), drawn from sentences in the C4 corpus. Negative examples were 1,000 tokens randomly selected from the same sentences, restricted to tokens whose decoded string consists entirely of alphabetic characters (no numbers, punctuation, or special characters), and excluding the preposition *up* itself and any token containing *up* as a substring. The validation set was drawn from the same source sentence pool (a non-overlapping subset), with token positions resolved per model's tokenizer; exact counts fall slightly below 1,000 for some models because *up* is not a standalone BPE token in every sentence under every tokenizer — see @tbl-exp1-splits.

The classifiers targeting OLMo-3 7B and the BabyLM models were trained on the **same underlying sentences**, but the frequency and predictability statistics used in the analyses were derived from the respective corpus for each model (Dolma v1.7 for OLMo-3 7B, queried via the infini-gram API; and the BabyLM training corpus for the BabyLM models). In other words, the training sentences were identical across models, but the tokenization and corpus statistics varied.

Classifier training and validation split sizes are shown in @tbl-exp1-splits (Appendix). The test set comprised verb–*up* phrasal verb types (e.g., *pick up*, *set up*) with at least 20 occurrences in the corpus; up to 20 sentences were sampled per type. Corpus frequency and predictability ($P(\text{up} \mid V)$) were derived from the corpus matched to each model's training distribution. Because the BabyLM corpus is smaller than Dolma, fewer V+up types attain a valid (non-zero) predictability estimate under BabyLM statistics; only types with a valid predictability estimate are included in the analyses. Full item-level statistics are reported in @tbl-test-items (Appendix).

For OLMo-3 7B, 3,927 unique V+*up* types were included, with a median corpus frequency of 92 (range: 20–390,086) and a median logit(predictability) of $-5.38$ (range: $-10.42$–$3.87$). The BabyLM models shared 1,039 items; these items had a higher median frequency (920) and higher median logit(predictability) ($-3.40$) under BabyLM-corpus statistics, reflecting the smaller and more concentrated vocabulary of the BabyLM training data.

## Joint model — predicted surfaces (linear)

```{r}
#| label: fig-exp1-joint-surfaces
#| fig-cap: "Predicted logit as a function of frequency (left) and predictability (right) from the linear joint brms model, at the first and final layers, for OLMo-3 7B (UP independently). Each line slices the joint surface at the mean of the other predictor."
plot_marginal_surfaces(surface_exp1_joint_olmo %>%
  mutate(label = if_else(str_detect(label, "first"), "First layer", "Final layer")))
```

```{r}
#| label: fig-exp1-joint-surfaces-blm
#| fig-cap: "Predicted logit as a function of frequency (left) and predictability (right) from the linear joint brms model at the first (top) and final (bottom) layers, for BabyLM models (UP independently). Each line slices the joint surface at the mean of the other predictor."
p1 = plot_marginal_surfaces(blm_strip(surface_exp1_joint_blm, "first")) + labs(title = "First layer")
p2 = plot_marginal_surfaces(blm_strip(surface_exp1_joint_blm, "final")) + labs(title = "Final layer")
(p1 / p2) + plot_layout(guides = "collect") & theme(legend.position = "right")
```

# Experiment 2: UP as Subword

## Items

The training set for this experiment combined two types of positive example: 1,000 occurrences of standalone *up* (identical to Experiment 1) and 1,000 occurrences of *up* embedded within a larger word (e.g., *update*, *upon*, *setup*). Critically, the subword positives were restricted to **unique word types**: each up-containing word form contributed exactly one instance, so the classifier could not learn to recognise a high-frequency form like *setup* from repeated exposure. Negative examples (1,000 drawn from each sentence pool) were tokens consisting entirely of alphabetic characters that did not contain *up* as a substring. The validation set was constructed by the same procedure (targeting 1,000 positive, 1,000 negative per sentence pool; exact counts vary by tokenizer — see @tbl-exp2-splits). As in Experiment 1, OLMo-3 7B and BabyLM models used the same underlying sentences with their own tokenizers, and BabyLM models used BabyLM-corpus statistics for frequency and predictability. The test set was identical to Experiment 1: all verb–*up* phrasal verb types with at least 20 corpus occurrences, sampled at up to 20 sentences per type.

Classifier training and validation split sizes are shown in @tbl-exp2-splits (Appendix). The OLMo training set contained 1,000 unique up-within-word types; BabyLM models similarly contributed 1,000 unique types, with the exact set varying by tokenizer. Validation-set sizes differ slightly across models because some sentences could not be resolved to a valid token position under each model's tokenizer. The test set is identical to Experiment 1; item-level statistics are shown in @tbl-test-items.

## Joint model — predicted surfaces (linear)

```{r}
#| label: fig-exp2-joint-surfaces
#| fig-cap: "Predicted logit as a function of frequency (left) and predictability (right) from the linear joint brms model, at the first and final layers, for OLMo-3 7B (UP as subword). Each line slices the joint surface at the mean of the other predictor."
plot_marginal_surfaces(surface_exp2_joint_olmo %>%
  mutate(label = if_else(str_detect(label, "first"), "First layer", "Final layer")))
```

```{r}
#| label: fig-exp2-joint-surfaces-blm
#| fig-cap: "Predicted logit as a function of frequency (left) and predictability (right) from the linear joint brms model at the first (top) and final (bottom) layers, for BabyLM models (UP as subword). Each line slices the joint surface at the mean of the other predictor."
p1 = plot_marginal_surfaces(blm_strip(surface_exp2_joint_blm, "first")) + labs(title = "First layer")
p2 = plot_marginal_surfaces(blm_strip(surface_exp2_joint_blm, "final")) + labs(title = "Final layer")
(p1 / p2) + plot_layout(guides = "collect") & theme(legend.position = "right")
```

# Experiment 3: Whisper (ASR Model)

## Items

The Whisper experiment used spoken language data drawn from a mixed-source audio metadata file (`up-audio-metadata.csv`) containing 224,118 timestamped speech segments in which *up* occurred. Word-level timestamps within each segment were obtained using WhisperX forced alignment. Each occurrence of *up* was classified by running spaCy part-of-speech tagging on the full segment transcript: if the token immediately preceding *up* was tagged as a VERB, the instance was labelled *V+up* (e.g., *pick up*, *clean up*); otherwise it was labelled *word_up* (e.g., *it up*, *them up*). This classification procedure mirrors the one used to identify standalone *up* in Experiments 1 and 2. Of the 224,118 segments, 58,751 were classified as *word_up* and 165,367 as *V+up*, spanning 4,161 unique V+up types.

The **training and validation sets** used 1,000 *word_up* instances each as positive examples, paired with 1,000 randomly selected non-*up* words from the same segment as negative examples. This mirrors the standalone-*up* training design of Experiments 1 and 2. Negative words were restricted to those whose string consists entirely of alphabetic characters (matching the `token.isalpha()` criterion used in Experiments 1 and 2), with a minimum duration of 20 ms. Training and validation segments were non-overlapping.

The **test set** comprised V+up phrasal verb types with at least 5 occurrences in the audio dataset, yielding 1,466 qualifying types, with up to 20 instances per type (median: 16 sentences per type; range: 5–20). The minimum-occurrence threshold is lower than in Experiments 1 and 2 (which required at least 20 corpus sentences per type), reflecting the sparser coverage of audio relative to written text. Corpus frequency and predictability values were drawn from the C4-derived lookup table used in Experiments 1 and 2. Classifier training and validation split sizes are shown in @tbl-exp3-splits (Appendix) and full item-level statistics are shown in @tbl-test-items (Appendix). After filtering to types with a valid C4 predictability estimate, 1,397 unique V+*up* types were retained for both encoder and decoder, with a median corpus frequency of 683 (range: 20–390,086) and a median logit(predictability) of $-4.05$ (range: $-10.42$–$3.87$).

## Joint model — predicted surfaces (linear)

```{r}
#| label: fig-exp3-joint-surfaces
#| fig-cap: "Predicted logit as a function of frequency (left) and predictability (right) from the linear joint brms model at the first (top) and final (bottom) layers, for Whisper encoder and decoder. Each line slices the joint surface at the mean of the other predictor."
p1 = plot_marginal_surfaces(wh_strip(surface_exp3_joint_wh, "first")) + labs(title = "First layer")
p2 = plot_marginal_surfaces(wh_strip(surface_exp3_joint_wh, "final")) + labs(title = "Final layer")
(p1 / p2) + plot_layout(guides = "collect") & theme(legend.position = "right")
```

```{=latex}
\clearpage
\bibliography{references}
\clearpage
```

\onecolumn

# Appendix {.appendix}

## BabyLM Model Hyperparameters

```{r}
#| label: tbl-hyperparams
#| tbl-pos: "H"
#| tbl-cap: "Hyperparameters for the three BabyLM models. All models share a BPE tokenizer (vocabulary size 8,192) trained on the BabyLM corpus. Training used two NVIDIA A100 80GB GPUs with distributed data parallelism."
tibble::tribble(
  ~Parameter,                   ~`OPT-125M`,           ~`OPT-350M`,           ~`OPT-1.3B`,
  "Hidden size",                 "768",                 "1,024",               "2,048",
  "Attention heads",             "12",                  "16",                  "32",
  "Layers",                      "12",                  "24",                  "24",
  "FFN intermediate size",       "3,072",               "4,096",               "8,192",
  "Vocabulary size",             "8,192",               "8,192",               "8,192",
  "Context length (tokens)",     "1,024",               "1,024",               "1,024",
  "Batch size (per device)",     "400",                 "200",                 "100",
  "Learning rate",               "$3\\times10^{-4}$",   "$1\\times10^{-4}$",   "$1\\times10^{-4}$",
  "Warmup steps",                "366",                 "732",                 "1,465",
  "Training epochs",             "20",                  "20",                  "20",
  "Optimizer",                   "AdamW (fused)",        "AdamW (fused)",        "AdamW (fused)",
  "Precision",                   "bfloat16",            "bfloat16",            "bfloat16"
) |>
  knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE,
               linesep = "", escape = FALSE)
```

\clearpage

## BabyLM Model Convergence

All three models trained to completion over 20 epochs, with training loss decreasing monotonically throughout (@fig-train-loss). Final gradient norms were small for all models (0.16, 0.24, and 0.38 for the 125M, 350M, and 1.3B models, respectively), and the learning rate decayed to zero at the end of training, consistent with full convergence under the scheduled warmup-then-decay regime. No signs of loss divergence or instability were observed.

The 1.3B model achieves the lowest training loss (2.55) and validation perplexity (15.18), as expected given its greater capacity. The 350M model, however, shows slightly elevated training loss (2.63) and validation perplexity (18.86) relative to the 125M model (training loss 2.55; validation perplexity 16.81). This is likely attributable to the lower learning rate applied to the 350M model ($1 \times 10^{-4}$ vs. $3 \times 10^{-4}$ for the 125M), which, combined with a relatively modest dataset size of 150M tokens, may have produced slower effective convergence. The effect is small and does not affect our qualitative comparisons, but we note it here for completeness.

```{r}
#| label: fig-train-loss
#| fig-pos: "H"
#| fig-cap: "Training loss over 20 epochs for each BabyLM model. Loss is logged every 10 gradient steps."
babylm_train_log |>
  mutate(model = factor(model, levels = c("OPT-125M", "OPT-350M", "OPT-1.3B"))) |>
  ggplot(aes(x = epoch, y = loss, color = model)) +
  geom_line(linewidth = 0.4, alpha = 0.9) +
  scale_color_manual(values = uo_colors) +
  labs(x = "Epoch", y = "Training loss", color = NULL) +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
#| label: tbl-val-perplexity
#| tbl-pos: "H"
#| tbl-cap: "Final validation loss and perplexity for each BabyLM model, evaluated on the BabyLM development set after 20 epochs of training."
tibble::tribble(
  ~Model,       ~`Val. loss`, ~Perplexity,
  "OPT-125M",    2.82,         16.81,
  "OPT-350M",    2.94,         18.86,
  "OPT-1.3B",    2.72,         15.18
) |>
  knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

```{r}
#| label: fig-val-perplexity
#| fig-pos: "H"
#| fig-cap: "Final validation perplexity for each BabyLM model after 20 epochs of training on the BabyLM corpus. Lower perplexity indicates better fit to the validation data."
tibble::tribble(
  ~Model,       ~Perplexity,
  "OPT-125M",   16.81,
  "OPT-350M",   18.86,
  "OPT-1.3B",   15.18
) |>
  mutate(Model = factor(Model, levels = c("OPT-125M", "OPT-350M", "OPT-1.3B"))) |>
  ggplot(aes(x = Model, y = Perplexity, fill = Model)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", Perplexity)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = uo_colors) +
  labs(x = NULL, y = "Validation perplexity") +
  ylim(0, 23) +
  theme_minimal()
```

\clearpage

## Test Set Items

```{r}
#| label: tbl-test-items
#| tbl-pos: "H"
#| tbl-cap: "Test set statistics for all experiments. Only items with a valid (non-zero) predictability estimate are included. Frequency is the raw corpus count; predictability is logit(predictability). BabyLM models (125M, 350M, 1.3B) share identical items and corpus statistics. Whisper encoder and decoder evaluate the same audio items with Dolma-derived frequency counts."
bind_rows(
  olmo_items %>% mutate(Model = "OLMo-3 7B"),
  blm_items  %>% mutate(Model = "BabyLM"),
  wh_items_by_comp %>% filter(component == "encoder") %>% mutate(Model = "Whisper encoder"),
  wh_items_by_comp %>% filter(component == "decoder") %>% mutate(Model = "Whisper decoder")
) %>%
  group_by(Model) %>%
  summarise(
    `N items`      = n(),
    `Med. freq`    = as.integer(median(frequency)),
    `Freq range`   = sprintf("%s--%s",
                      format(min(frequency), big.mark = ","),
                      format(max(frequency), big.mark = ",")),
    `Med. predic.` = round(median(log_predic, na.rm = TRUE), 2),
    `Predic. range`= sprintf("%.2f--%.2f", min(log_predic, na.rm = TRUE), max(log_predic, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(Model = factor(Model, levels = c("OLMo-3 7B", "BabyLM",
                                           "Whisper encoder", "Whisper decoder"))) %>%
  arrange(Model) %>%
  knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

## Experiment 1: UP Independently

```{r}
#| label: tbl-exp1-splits
#| tbl-pos: "H"
#| tbl-cap: "Classifier training and validation split sizes for Experiment 1 (UP independently). Positive (pos) = standalone *up*; negative (neg) = other alphabetic token from the same sentence."
tibble::tribble(
  ~Model,         ~`Train pos`, ~`Train neg`, ~`Val pos`, ~`Val neg`,
  "OLMo-3 7B",   1000L,        1000L,        992L,       1000L,
  "BabyLM 125M", 1000L,        1000L,        999L,       1000L,
  "BabyLM 350M", 1000L,        1000L,        999L,       1000L,
  "BabyLM 1.3B", 1000L,        1000L,        916L,       1000L
) %>% knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE, format.args = list(big.mark = ","))
```

```{r}
#| label: tbl-exp1-joint-first
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the first layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp1_joint_first), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

```{r}
#| label: tbl-exp1-joint-final
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the final layer for UP independently. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp1_joint_final), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

\clearpage

## Experiment 2: UP as Subword

```{r}
#| label: tbl-exp2-splits
#| tbl-pos: "H"
#| tbl-cap: "Classifier training and validation split sizes for Experiment 2 (UP as subword). Positive examples: 1,000 standalone *up* + 1,000 unique up-within-word types. Negative examples: 1,000 from each sentence pool. Val sizes vary by model due to tokenizer coverage."
tibble::tribble(
  ~Model,         ~`Train standalone`, ~`Train subword`, ~`Train neg`, ~`Val rows`,
  "OLMo-3 7B",   1000L,               1000L,            2000L,        3868L,
  "BabyLM 125M", 1000L,               1000L,            2000L,        3582L,
  "BabyLM 350M", 1000L,               1000L,            2000L,        3582L,
  "BabyLM 1.3B", 1000L,               1000L,            2000L,        3773L
) %>% knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE, format.args = list(big.mark = ","))
```

```{r}
#| label: tbl-exp2-joint-first
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the first layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp2_joint_first), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

```{r}
#| label: tbl-exp2-joint-final
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the final layer for UP as subword. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp2_joint_final), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

\clearpage

## Experiment 3: Whisper

```{r}
#| label: tbl-exp3-splits
#| tbl-pos: "H"
#| tbl-cap: "Classifier training and validation split sizes for Experiment 3 (Whisper). Positive examples are *word_up* instances; negatives are random non-*up* alphabetic tokens from the same segment."
wh_splits %>%
  select(Component, `Train pos`, `Train neg`, `Val pos`, `Val neg`) %>%
  knitr::kable(row.names = FALSE, format = "latex", booktabs = TRUE, format.args = list(big.mark = ","))
```

```{r}
#| label: tbl-exp3-joint-first
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the first layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp3_joint_first), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

```{r}
#| label: tbl-exp3-joint-final
#| tbl-pos: "H"
#| tbl-cap: "Joint frequency × predictability model at the final layer for Whisper encoder and decoder. Estimates are from Bayesian linear mixed-effects models (brms); % > 0 indicates the percentage of posterior samples with a positive estimate."
knitr::kable(clean_brms_table(tbl_exp3_joint_final), row.names = FALSE, format = "latex", booktabs = TRUE, linesep = "")
```

\clearpage
