---
title: "analysis-script — Whisper-small (Encoder vs Decoder)"
author: "Zachary Houghton"
date: "2026-03-07"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(brms)
library(mgcv)
library(tidybayes)
library(patchwork)
library(viridis)
library(plotly)
```

# Analysis — Whisper-small (Encoder vs Decoder)

## Load data

The encoder and decoder each have 12 layers (0–11). Both are loaded and
combined with a `component` factor for joint modelling.

```{r}
encoder = read_csv('../../Data/whisper/encoder/all_layers_results.csv') %>%
  mutate(component = "encoder", log_freq = log(frequency), verb_up = factor(verb_up))

decoder = read_csv('../../Data/whisper/decoder/all_layers_results.csv') %>%
  mutate(component = "decoder", log_freq = log(frequency), verb_up = factor(verb_up))

whisper = bind_rows(encoder, decoder) %>%
  mutate(component = factor(component, levels = c("encoder", "decoder")))

# Final-layer (layer 11) subsets
enc_final = encoder %>% filter(layer == 11)
dec_final = decoder %>% filter(layer == 11)
whisper_final = whisper %>% filter(layer == 11)
```

## Effect of frequency at the final layer — by component (brms)
```{r}
priors = c(
  prior(normal(0, 1), class = 'b')
)

model_freq_final = brm(
  formula = logit ~ log_freq * component + (1 | verb_up),
  data    = whisper_final,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_final",
  seed    = 964
)
summary(model_freq_final)
```

## Effect of frequency at the first layer — by component (brms)
```{r}
whisper_first = whisper %>% filter(layer == 0)

model_freq_first = brm(
  formula = logit ~ log_freq * component + (1 | verb_up),
  data    = whisper_first,
  prior   = priors,
  iter    = 6000,
  warmup  = 3000,
  chains  = 4,
  cores   = 4,
  file    = "model_freq_first",
  seed    = 964
)
summary(model_freq_first)
```

## Effect of frequency at the final layer — by component (bam — non-linear)
```{r}
set.seed(964)
if (file.exists("bam_freq_final.rds")) {
  bam_freq_final = readRDS("bam_freq_final.rds")
} else {
  bam_freq_final = bam(
    formula  = logit ~ component + s(log_freq, by = component) + s(verb_up, bs = 're'),
    data     = whisper_final,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(bam_freq_final, "bam_freq_final.rds")
}
summary(bam_freq_final)
```

## Effect of frequency across all layers — by component (mgcv)

Both encoder and decoder have 12 layers (0–11), so no normalization is needed.

```{r}
set.seed(964)
if (file.exists("model_freq_layer.rds")) {
  model_freq_layer = readRDS("model_freq_layer.rds")
} else {
  model_freq_layer = bam(
    formula  = logit ~ component + te(log_freq, layer, by = component) + s(verb_up, bs = 're'),
    data     = whisper,
    method   = 'fREML',
    discrete = TRUE
  )
  saveRDS(model_freq_layer, "model_freq_layer.rds")
}
summary(model_freq_layer)
```

## Plots

### Final and first layer — frequency effect by component (brms)
```{r}
plot_layer_by_component = function(model_fit, data, title) {
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 100)

  newdata = expand.grid(
    log_freq  = log_freq_seq,
    component = levels(data$component)
  ) %>% mutate(verb_up = NA)

  epred = add_epred_draws(newdata, model_fit, re_formula = NA, ndraws = 500)

  epred %>%
    ggplot(aes(x = log_freq, y = .epred, color = component, fill = component)) +
    stat_lineribbon(.width = 0.95, alpha = 0.2) +
    stat_summary(
      data  = data,
      aes(x = log_freq, y = logit, color = component),
      fun   = mean, geom = "point", size = 1, alpha = 0.5
    ) +
    labs(
      title = title,
      x     = "log(frequency)",
      y     = "Predicted logit",
      fill  = "Component", color = "Component"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 11, face = "bold", margin = margin(b = 8)),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}

p_final = plot_layer_by_component(
  model_freq_final, whisper_final,
  "Whisper-small: frequency effect at final layer (11) — encoder vs decoder"
)
p_first = plot_layer_by_component(
  model_freq_first, whisper_first,
  "Whisper-small: frequency effect at first layer (0) — encoder vs decoder"
)

p_first / p_final
```

### Across layers — frequency effect by component (mgcv)
```{r}
plot_freq_by_layer = function(model_fit, data, title) {
  layer_seq    = 0:11
  log_freq_seq = seq(min(data$log_freq), max(data$log_freq), length.out = 50)
  log_freq_lo  = quantile(data$log_freq, 0.1)
  log_freq_hi  = quantile(data$log_freq, 0.9)

  # ── Spaghetti ──────────────────────────────────────────────────────────────
  newdata_sp = expand.grid(
    layer     = layer_seq,
    log_freq  = log_freq_seq,
    component = levels(data$component)
  ) %>% mutate(verb_up = data$verb_up[1])

  preds_sp = predict(model_fit, newdata = newdata_sp, se.fit = TRUE,
                     exclude = 's(verb_up)')
  newdata_sp$fit = preds_sp$fit

  p_sp = newdata_sp %>%
    ggplot(aes(x = log_freq, y = fit, color = layer, group = interaction(component, layer))) +
    geom_line(linewidth = 0.5, alpha = 0.7) +
    facet_wrap(~component) +
    scale_color_viridis_c(option = "plasma", name = "Layer") +
    labs(
      title = paste0(title, ": frequency effect by layer — encoder vs decoder"),
      x = "log(frequency)", y = "Predicted logit"
    ) +
    theme_minimal()

  # ── Difference (high - low freq) across layers ─────────────────────────────
  newdata_diff = expand.grid(
    layer     = layer_seq,
    component = levels(data$component),
    freq_level = c("low", "high")
  ) %>%
    mutate(
      log_freq = ifelse(freq_level == "low", log_freq_lo, log_freq_hi),
      verb_up  = data$verb_up[1]
    )

  preds_diff = predict(model_fit, newdata = newdata_diff, se.fit = TRUE,
                       exclude = 's(verb_up)')
  newdata_diff$fit = preds_diff$fit
  newdata_diff$se  = preds_diff$se.fit

  p_diff = newdata_diff %>%
    select(layer, component, freq_level, fit, se) %>%
    pivot_wider(names_from = freq_level, values_from = c(fit, se)) %>%
    mutate(
      diff    = fit_high - fit_low,
      se_diff = sqrt(se_high^2 + se_low^2),
      lo95    = diff - 1.96 * se_diff,
      hi95    = diff + 1.96 * se_diff
    ) %>%
    ggplot(aes(x = layer, y = diff, color = component, fill = component)) +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.2, color = NA) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(title, ": frequency effect (high - low) across layers"),
      x = "Layer", y = "Difference in predicted logit (90th - 10th pct freq)",
      color = "Component", fill = "Component"
    ) +
    theme_minimal()

  # ── Heatmap ────────────────────────────────────────────────────────────────
  newdata_heat = expand.grid(layer = layer_seq, log_freq = log_freq_seq,
                              component = levels(data$component)) %>%
    mutate(verb_up = data$verb_up[1])
  preds_heat = predict(model_fit, newdata = newdata_heat, se.fit = FALSE,
                        exclude = 's(verb_up)')
  newdata_heat$fit = preds_heat

  p_heat = newdata_heat %>%
    ggplot(aes(x = layer, y = log_freq, fill = fit)) +
    geom_tile() +
    facet_wrap(~component) +
    scale_fill_viridis_c(option = "magma", name = "Predicted\nlogit") +
    labs(title = paste0(title, ": predicted logit surface"),
         x = "Layer", y = "log(frequency)") +
    theme_minimal()

  list(spaghetti = p_sp, diff = p_diff, heatmap = p_heat)
}

plots = plot_freq_by_layer(model_freq_layer, whisper, "Whisper-small")

plots$spaghetti
plots$diff
plots$heatmap
```

### Encoder vs decoder comparison — val accuracy across layers
```{r}
# Load layer metadata from JSON files
read_layer_meta = function(path, component) {
  meta = jsonlite::fromJSON(path)
  tibble(
    layer       = sapply(meta$layers, `[[`, "layer"),
    component   = component,
    val_acc     = sapply(meta$layers, `[[`, "val_acc"),
    val_up_acc  = sapply(meta$layers, `[[`, "val_up_acc"),
    val_oth_acc = sapply(meta$layers, `[[`, "val_other_acc"),
    cv_mean     = sapply(meta$layers, `[[`, "cv_mean"),
  )
}

meta_enc = read_layer_meta('../../Data/whisper/encoder/layer_metadata.json', 'encoder')
meta_dec = read_layer_meta('../../Data/whisper/decoder/layer_metadata.json', 'decoder')
meta = bind_rows(meta_enc, meta_dec) %>%
  mutate(component = factor(component, levels = c("encoder", "decoder")))

meta %>%
  ggplot(aes(x = layer, y = val_acc, color = component)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("encoder" = "#1a9850", "decoder" = "#d73027")) +
  labs(
    title = "Whisper-small: classifier validation accuracy across layers",
    x = "Layer", y = "Validation accuracy",
    color = "Component"
  ) +
  theme_minimal(base_size = 13)
```

---

## Encoder vs decoder — direct logit comparison at each layer

This shows whether the encoder or decoder representations are more sensitive
to V+up compositionality at each layer.

```{r}
# Mean logit per V+up type, per layer, per component
summary_by_layer = whisper %>%
  group_by(component, layer, verb_up, frequency) %>%
  summarise(mean_logit = mean(logit), .groups = "drop")

# Spearman correlation between log(freq) and mean logit, per layer per component
cor_by_layer = summary_by_layer %>%
  group_by(component, layer) %>%
  summarise(
    spearman_r = cor(log(frequency), mean_logit, method = "spearman"),
    .groups = "drop"
  )

cor_by_layer %>%
  ggplot(aes(x = layer, y = spearman_r, color = component)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("encoder" = "#1a9850", "decoder" = "#d73027")) +
  labs(
    title = "Spearman r(log freq, mean logit) across layers — encoder vs decoder",
    subtitle = "Negative: high-freq V+up types look less like standalone 'up'",
    x = "Layer", y = "Spearman r",
    color = "Component"
  ) +
  theme_minimal(base_size = 13)
```

---

## Most and least compositional V+up types (final layer)
```{r}
plot_top_bottom = function(data, component_name, layer_num = 11) {
  d = data %>%
    filter(component == component_name, layer == layer_num) %>%
    group_by(verb_up, frequency) %>%
    summarise(mean_logit = mean(logit), .groups = "drop")

  n_show = 20
  bar_df = bind_rows(
    d %>% slice_max(mean_logit, n = n_show),
    d %>% slice_min(mean_logit, n = n_show)
  ) %>%
    distinct(verb_up, .keep_all = TRUE) %>%
    arrange(mean_logit)

  colors = ifelse(bar_df$mean_logit < 0, "#d73027", "#1a9850")

  bar_df %>%
    ggplot(aes(x = mean_logit, y = reorder(verb_up, mean_logit))) +
    geom_col(fill = colors, alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    labs(
      title = paste0(component_name, " (layer ", layer_num, "): most & least compositional V+up"),
      x = "Mean logit (higher = more standalone-'up'-like)", y = NULL
    ) +
    theme_minimal(base_size = 11)
}

plot_top_bottom(whisper, "encoder") /
plot_top_bottom(whisper, "decoder")
```
